---
title: 'Project 1 Report'
author: |
  |
  |
  | Analyst: Isabelle Meredith
  | Investigators: Nichole Carlson
  | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    highlight: espresso
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
---

# Introduction

<!-- In order to provide context for the analysis both to the investigator and other statisticians, briefly describe the project itself. This would include a high-level overview of the study design including primary and secondary outcomes and the study population or datasets you received. Then, in order to link the overall project goals to the specific statistical analyses, describe, as specifically as possible, a clearly defined research question. For most projects, this will be some sort of research hypothesis or statistical hypothesis. This section should have enough detail to allow a person to see how the data (described below) can be employed to address the overall goals of the project (described above).  Also include a statement of what this report addressed. -->

<!-- This can basically be the data analysis plan with any changes suggested to you from the investigator from that report. -->

The data for this project come from the prospective cohort study, Multicenter AIDS Cohort Study, which investigates the natural and treated courses of HIV-1 in gay and bisexual men in four major US cities. This study focuses on the subset of the data where participants were prescribed highly active antiretroviral treatment (HAART), the standard treatment for HIV-1. Measures of physical and mental health were taken at baseline before treatment and annually each year after for eight years. The goal of this secondary analysis is to examine whether there is a difference in treatment efficacy at two years after treatment initiation between subjects who were using hard drugs at baseline and subjects who were not. The outcomes of interest in this analysis are viral load, CD4+ T cell count, an aggregate physical quality of life score, and an aggregate mental quality of life score. 

The purpose of this report is to develop the analysis plan to investigate the following hypothesis:

1. Hypothesis: Treatment response to HAART as measured by viral load, CD4+ T cell count, an aggregate physical quality of life score, and an aggregate mental quality of life score is lower in subjects who were hard drug users at baseline than in subjects who were not.

# Methods

<!-- Things to consider addressing. This list is more comprehensive than what is the in the DataAnalyisPlan_Report. -->

<!-- *	Data quality –elements of data collection and pre-processing that could influence data quality -->

<!-- *	Data quality checking methods (looking for outliers, illogical results, etc. -->
<!--   +	If missing data were imputed, then methods used and actions taken to ensure that bias was not introduced and that variance was not compromised should be described -->
<!--   +	Description of methods which have been adopted to minimize the amount of missing data -->

<!-- *	Exploratory methods/exploratory analysis summary -->
<!--   +	Approaches used (graphical, hypothesis tests, expert opinions) -->
<!--   +	Data transformation and manipulation -->

<!-- *	The statistical methods used to analyze the outcomes listed in the review of project -->
<!--   +	Could write it generally (i.e., not in reference to specific outcomes) -->
<!--   + List crude analyses, if applicable. -->
<!--   +	List any variables you are adjusting for in your models (precision vs. confounding vs. face-validity) -->
<!--   +	Describe any variable selection methods used, if applicable. -->
<!--   + Modeling assumptions -->
<!--   + Assumption checks/sensitivity analyses -->

<!-- * Decision criteria (p-value threshold) and discussion on multiple comparisons, if applicable. -->

<!-- *	Software used for analyses -->

<!--## Data Cleaning


List any data cleaning that will be required (or has already been done). This would include new representation of variables (collapsing groups) and observations that had to be removed, etc.  

Example:

*Race was collapsed into white and non-white given small sample sizes in each of the non-white categories. Variable X was dichotomized as Large and Small given clinical thresholds [made up example]. Two individuals were removed from analysis given contaminated samples.  Four other individuals had implausible values. The analysis was completed with and without those four samples.*


## Data Analysis

For the project as a whole, and for each research question, aim, or hypothesis, describe the proposed analytic method and any necessary details for its implementation. At a minimum, this should address the hypotheses mentioned above. For further detail, describe any follow up or secondary analyses, any assessments of assumptions, and sensitivity or subgroup analyses. Do not forget to include treatment of missing data. This section should have enough detail that any statistician could recreate the analysis using any statistical software and the source data. 

Example:

*A univariable analysis between outcomes and exposure and outcomes and covariates/confounders will be performed using general linear regression. Assumptions of linearity, homoscedasticity, and normality will be confirmed using diagnostic plots. A complete case analysis will be used for each model. Covariates with a significant association with the outcome will be used for adjustment in a multivariable model. P values <0.05 will be considered significant. Baseline FEV, gender, and age will be included in the adjusted model based on their biological associations with the outcome. Since participants were randomly assigned, there should be no relationship between possible covariates and exposure, this will be confirmed using χ2/fisher’s exact for categorical covariates and ANOVA for continuous covariates.*

-->

## Data Cleaning

The initial dataset  of 715 subjects was filtered to include only subjects that had values for all outcomes at baseline and year 2. 476 subjects had full outcome data at baseline and year 2. Viral load was base 10 log transformed as viral loads can vary from tenths of a copy per mL to millions of copies per mL. In order to have categories with sufficient counts to be meaningful, education level was collapsed into "college degree or higher" and "didn't finish college or lower". Similarly, race was collapsed into "white" and "other". Histograms were used to locate outliers and other potential data problems. 13 subjects had either negative BMIs or BMIs over 100. These values were considered data entry errors and the subjects were removed, as we performed a complete case analysis, leaving 463 subjects. However, the subjects with missing BMI are reported in table 1 for transparency.

## Data Analysis

The changes in each of the four outcomes between year zero and year two were considered the primary outcomes of interest. A univariable analysis between the outcomes and hard drug use at baseline was performed using general linear regression in both a frequentist and a Bayesian framework. The frequentist and Bayesian analyses were conducted separately. Assumptions of linearity, homoscedasticity, and normality were confirmed using diagnostic plots. A complete case analysis was used for each model. P values <0.05 were considered significant for the frequentist models. Leave one out cross-validation (Loo) was used to assess whether a model had meaningful improvement for the Bayesian models. The baseline value for the outcome of each model was included in the model as adjustment. Age, BMI, smoking, race, and education were included as covariates due to their relevance in outcomes in previous studies. 

The Bayesian models used non-informative normal priors with mean 0 and standard deviation 100 for each of the beta coefficients and a half normal with mean 0 and standard deviation 100 for the residual standard error. Each model had four chains with 40,000 iterations of which 10,000 were warm up. The Hamiltonian Monte Carlo No U-Turn Sampler (HMC-NUTS) was used. MCMC trace and density plots were examined to confirm successful convergence.

In order to assess whether adherence to treatment is a mediator in this analysis, a version of each model was then created with and without the adherence level. The difference in the coefficient for hard drug use between the model with adherence and the model without adherence was calculated. This was intended as a preliminary analysis for a mediation effect, and based on whether preliminary findings are suggestive, a full mediation analysis will be considered for future work.

All models were run using R version 4.4.3. The brms package was used to estimate Bayesian models.

# Results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(gtsummary)
library(kableExtra)

# create Table 1
dfHIV <- read.csv("./DataRaw/hiv_6624_final.csv") %>% 
  filter(years == 0 | years == 2) %>% 
  select(newid, AGG_MENT, AGG_PHYS, BMI, SMOKE, LEU3N, VLOAD, RACE, EDUCBAS, age, ADH, years, hard_drugs) %>% 
  pivot_wider(names_from = years, values_from = c(AGG_MENT, AGG_PHYS, BMI, SMOKE, LEU3N, VLOAD, RACE, EDUCBAS, age, ADH, hard_drugs)) %>%
  filter(!(is.na(AGG_MENT_0) | is.na(AGG_MENT_2) | is.na(AGG_PHYS_0)| is.na(AGG_PHYS_2) | is.na(LEU3N_0) | is.na(LEU3N_2) | is.na(VLOAD_0) | is.na(VLOAD_2))) %>%
  select(-c(BMI_2, SMOKE_2, RACE_2, EDUCBAS_2, age_2, ADH_0, hard_drugs_2)) %>%
  mutate(BMI_0 = ifelse((BMI_0 > 100) | (BMI_0 < 0), NA, BMI_0))

dfTableOne <- dfHIV %>% 
  mutate(EDUCBAS_0 = case_match(EDUCBAS_0, c(1, 2, 3, 4) ~ "Less than a college degree",
                                c(5, 6, 7) ~ "College degree or higher"),
         RACE_0 = case_match(RACE_0, 1 ~ "White, non-Hispanic",
                             c(2, 3, 4, 7, 8) ~ "Other"),
         SMOKE_0 = case_match(SMOKE_0, c(1,2) ~ "Not current smoker",
                              3 ~ "Current smoker"),
         ADH_2 = case_match(ADH_2, c(1,2) ~ "95-100% Adherence",
                            c(3,4) ~ "<95% Adherence"),
         hard_drugs_0 = if_else(hard_drugs_0 == 0, "No hard drugs", "Hard drugs"),
         logVLOAD_0 = log(VLOAD_0, base=10),
         logVLOAD_2 = log(VLOAD_2, base=10)) %>% 
  select(-c(VLOAD_0, VLOAD_2, newid)) %>%
  rename("Baseline Mental Score" = AGG_MENT_0,
         "Year 2 Mental Score" = AGG_MENT_2,
         "Baseline Physical Score" = AGG_PHYS_0,
         "Year 2 Physical Score" = AGG_PHYS_2,
         "Baseline Log Viral Load" = logVLOAD_0,
         "Year 2 Log Viral Load" = logVLOAD_2,
         "Baseline CD4+ T Count" = LEU3N_0,
         "Year 2 CD4+ T Count" = LEU3N_2,
         "Baseline BMI" = BMI_0,
         "Baseline Education" = EDUCBAS_0,
         "Smoking Status at Baseline" = SMOKE_0,
         "Age at Baseline" = age_0,
         "Adherence to Treatment" = ADH_2,
         "Hard Drugs" = hard_drugs_0,
         "Race" = RACE_0)
  
dfTableOne %>% tbl_summary(by="Hard Drugs") %>%
    as_kable_extra(booktabs = TRUE) %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down"` argument here.
  kableExtra::kable_styling(font_size = 10)
```

Subject demographics are reported in table 1. 

## Effect of Hard Drug Use

The effect of hard drug use is summarized for both the frequentist and Bayesian approaches in table 2.

```{r, echo=FALSE}
dfResults <- read.csv("./StoredResults/ResultsTable.csv")
kable(dfResults, col.names = c("Variable", "Frequentist Estimate", "95% Confidence Interval", "p Value", "Bayesian Estimate", "95% Credible Interval", "Posterior Probability"))
```

### Frequentist

Hard drug use was associated with a nonsignificant 0.29 point decrease in aggregate mental health score (95% CI: -3.75, 3.17, p = 0.870).  It was associated with a significant 3.34 point decrease in aggregate physical health score (95% CI: -6.06, -0.61, p = 0.017).  It was also associated with a significantly lower CD4+ T cell count, with 163.88	fewer cells (95% CI: -226.35, -101.41, p < 0.001). Finally, hard drug use was associated with a nonsignificant decrease of 0.06 log(viral load) (95% CI: (-0.46, 0.34), p = 0.770).

### Bayesian 

Hard drug use was associated with a 0.29 point decrease in aggregate mental health score. There is a 95% probability that the effect of hard drug use is between a 3.75 point decrease and a 3.14 increase in aggregate mental health score. The posterior probability that hard drug use results in a clinically meaningful 2 point difference in score is 26.2%. TALK ABOUT LOOIC I GUESS

Hard drug use was associated with a 3.33 point decrease in aggregate physical health score. There is a 95% probability that the effect of hard drug use is between a 6.06 and 0.60 point decrease in aggregate physical health score. The posterior probability that hard drug use results in a clinically meaningful 2 point difference in score is 83.2%. TALK ABOUT LOOIC I GUESS

Hard drug use was associated with a 148.96 cell decrease in CD4+ T cell count. There is a 95% probability that the effect of hard drug use is between a 208.5 and 89.38 cell decrease in CD4+ T cell count. The posterior probability that hard drug use results in a clinically meaningful 50 cell difference is 99.9%. TALK ABOUT LOOIC I GUESS

Hard drug use was associated with a 0.06 decrease in log(viral load). There is a 95% probability that the effect of hard drug use is between a 0.46 point decrease and a 0.34 increase in log(viral load). The posterior probability that hard drug use results in a clinically meaningful 0.5 difference in log(viral load) is 1.9%.  TALK ABOUT LOOIC I GUESS

<!-- A one-year increase in age is associated with a 2.7% increase in PSA. There is a 95% -->
<!-- probability of that the age effect is between a 0.55% decrease to a 5.9% increase in -->
<!-- PSA. -->
<!-- In addition, the posterior probability that a one-year increase in age results in a -->
<!-- change in PSA of 5% or greater is 7.9%. -->
<!-- Finally, based on the LOO-IC, age does not improve model fit for PSA by a -->
<!-- meaningful amount ($\Delta$ LOO-IC = -0.2). -->

## Mediation by Adherence

The impact of adherence on the effect of hard drug use is summarized for both the frequentist and Bayesian approaches in table 3.

```{r, echo=FALSE}
dfResultsAdherence <- read.csv("./StoredResults/ResultsAdherenceTable.csv")
kable(dfResultsAdherence, col.names = c("Variable", "Frequentist Estimate (95% CI)", "Frequentist Estimate with Adherence (95% CI)", "Delta AIC", "Bayesian Estimate (95% CI)", "Bayesian Estimate with Adherence (95% CI)", "Delta LOO-IC"))
```

### Frequentist

### Bayesian


<!-- Tables and figures can come at the end of the report or be interleaved. -->

<!-- Table 1 is a table describing the dataset and often with separate summaries by the primary variable of interest (e.g., treatment group, disease/no-disease, other as defined by hypothesis). This table usually doesn't have p-values, but sometimes for space limitations in journals, will have p-values. Be careful about whether you are using standard deviations vs. standard errors. SD is used without p-values. SE with p-values. -->

<!-- Tables 2-X are tables of results.  I usually have a presentation of basic findings prior to adjustment next to findings with adjustment. How the tables are organized will be project specific. -->

<!-- Graphic/plot ideas: histograms by variable of interest. Bar chars of major findings with means and 95% CI and p-values. Fitted regression lines for regression based analysis. Color is fine for class. Remember to ask an investigator if they need black and white for publication. -->

<!-- Total 6 tables and figures.  Outside of class, more than this would be an appendix or up for discussion on what to keep.  The number of tables and figures is limited in nearly all journals. -->

<!-- First paragraph are descriptive statistics and graphs with a description of the patterns of importance in a table/figures describing the data. -->

<!-- ## Associations between primary outcome and treatment -->

<!-- Go on to describe the results from your crude analyses and then your adjusted analyses.  This should include a reference to a results table, differences in the outcome by treatment group with estimates, confidence intervals and p-values. You might then compare how these crude results change with adjustment for precision/confounding variables (your final models). -->

<!-- ## Additional subheaders as necessary to guide through a more complex analysis. -->



# Discussion and Summary

<!-- Here you give a big picture of the major findings.  This allows you to highlight things you found most striking and perhaps even downplay findings that may be more spurious (e.g. a random interaction that pops up significant). -->

<!-- You then can give discussion on the limitations of the work.  Missing data biases (if any). Small samplesize (lack of power perhaps--clincially large findings that were insignificant). Small non-clinically meaningful differences that were signficant. -->

<!-- Assumptions that may have been violated and suggestions for further investigation. -->

# Reproducibility

Code is available at https://github.com/isabellemeredith/BIOS6624/tree/main/Project1
Data was located at "/Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project1/DataRaw/hiv_6624_final.csv"
