---
title: 'Project 1 Report'
author: |
  |
  |
  | Analyst: Isabelle Meredith
  | Investigators: Nichole Carlson
  | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
output:
  pdf_document:
    toc: no
    toc_depth: '2'
  html_document:
    highlight: espresso
    number_sections: yes
    theme: yeti
    toc: no
    toc_depth: 2
    toc_float: yes
---

# Introduction

The data for this project come from the prospective cohort study, Multicenter AIDS Cohort Study, which investigates the natural and treated courses of HIV-1 in gay and bisexual men in four major US cities. This study focuses on the subset of the data where participants were prescribed highly active antiretroviral treatment (HAART), the standard treatment for HIV-1. Measures of physical and mental health were taken at baseline before treatment and annually each year after for eight years. The goal of this secondary analysis is to examine whether there is a difference in treatment efficacy at two years after treatment initiation between subjects who were using hard drugs at baseline and subjects who were not. The outcomes of interest in this analysis are viral load, CD4+ T cell count, an aggregate physical quality of life score, and an aggregate mental quality of life score. 

The purpose of this report is to develop the analysis plan to investigate the following hypothesis:

1. Hypothesis: Treatment response to HAART as measured by viral load, CD4+ T cell count, an aggregate physical quality of life score, and an aggregate mental quality of life score is lower in subjects who were hard drug users at baseline than in subjects who were not.

# Methods

## Data Cleaning

The initial dataset  of 715 subjects was filtered to include only subjects that had values for all outcomes at baseline and year 2. 476 subjects had full outcome data at baseline and year 2. Viral load was base 10 log transformed as viral loads can vary from tenths of a copy per mL to millions of copies per mL. In order to have categories with sufficient counts to be meaningful, education level was collapsed into "college degree or higher" and "didn't finish college or lower". Similarly, race was collapsed into "white" and "other". Histograms were used to locate outliers and other potential data problems. 13 subjects had either negative BMIs or BMIs over 100. These values were considered data entry errors and the subjects were removed, as we performed a complete case analysis, leaving 463 subjects. However, the subjects with missing BMI are reported in table 1 for transparency.

## Data Analysis

The changes in each of the four outcomes between year zero and year two were considered the primary outcomes of interest. A univariable analysis between the outcomes and hard drug use at baseline was performed using general linear regression in both a frequentist and a Bayesian framework. The frequentist and Bayesian analyses were conducted separately. Assumptions of linearity, homoscedasticity, and normality were confirmed using diagnostic plots. A complete case analysis was used for each model. P values <0.05 were considered significant for the frequentist models. Changes in Akaike information criterion (AIC) of 3 or more were considrered meaningful for frequentist models. Leave-one-out cross-validation information criterion (LOO-IC) was used to assess whether a model had meaningful improvement for the Bayesian models, with a change of 3 being considered meaningful. The baseline value for the outcome of each model was included in the model as adjustment. Age, BMI, smoking, race, and education were included as covariates due to their relevance in outcomes in previous studies. 

The Bayesian models used non-informative normal priors with mean 0 and standard deviation 100 for each of the beta coefficients and a half normal with mean 0 and standard deviation 100 for the residual standard error. Each model had four chains with 40,000 iterations of which 10,000 were warm up. The Hamiltonian Monte Carlo No U-Turn Sampler (HMC-NUTS) was used. MCMC trace and density plots were examined to confirm successful convergence.

In order to assess whether adherence to treatment is a mediator in this analysis, a version of each model was then created with and without the adherence level. The difference in the coefficient for hard drug use between the model with adherence and the model without adherence was calculated. This was intended as a preliminary analysis for a mediation effect, and based on whether preliminary findings are suggestive, a full mediation analysis will be considered for future work.

All models were run using R version 4.4.3. The brms package was used to estimate Bayesian models. All Bayesian models were given the seed 28 for reproducibility.

# Results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(gtsummary)
library(kableExtra)
library(ggplot2)

# create Table 1
dfHIV <- read.csv("./DataRaw/hiv_6624_final.csv") %>% 
  filter(years == 0 | years == 2) %>% 
  select(newid, AGG_MENT, AGG_PHYS, BMI, SMOKE, LEU3N, VLOAD, RACE, EDUCBAS, age, ADH, years, hard_drugs) %>% 
  pivot_wider(names_from = years, values_from = c(AGG_MENT, AGG_PHYS, BMI, SMOKE, LEU3N, VLOAD, RACE, EDUCBAS, age, ADH, hard_drugs)) %>%
  filter(!(is.na(AGG_MENT_0) | is.na(AGG_MENT_2) | is.na(AGG_PHYS_0)| is.na(AGG_PHYS_2) | is.na(LEU3N_0) | is.na(LEU3N_2) | is.na(VLOAD_0) | is.na(VLOAD_2))) %>%
  select(-c(BMI_2, SMOKE_2, RACE_2, EDUCBAS_2, age_2, ADH_0, hard_drugs_2)) %>%
  mutate(BMI_0 = ifelse((BMI_0 > 100) | (BMI_0 < 0), NA, BMI_0))

dfTableOne <- dfHIV %>% 
  mutate(EDUCBAS_0 = case_match(EDUCBAS_0, c(1, 2, 3, 4) ~ "Less than a college degree",
                                c(5, 6, 7) ~ "College degree or higher"),
         RACE_0 = case_match(RACE_0, 1 ~ "White, non-Hispanic",
                             c(2, 3, 4, 7, 8) ~ "Other"),
         SMOKE_0 = case_match(SMOKE_0, c(1,2) ~ "Not current smoker",
                              3 ~ "Current smoker"),
         ADH_2 = case_match(ADH_2, c(1,2) ~ "95-100% Adherence",
                            c(3,4) ~ "<95% Adherence"),
         hard_drugs_0 = if_else(hard_drugs_0 == 0, "No hard drugs", "Hard drugs"),
         logVLOAD_0 = log(VLOAD_0, base=10),
         logVLOAD_2 = log(VLOAD_2, base=10)) %>% 
  select(-c(VLOAD_0, VLOAD_2, newid)) %>%
  rename("Baseline Mental Score" = AGG_MENT_0,
         "Year 2 Mental Score" = AGG_MENT_2,
         "Baseline Physical Score" = AGG_PHYS_0,
         "Year 2 Physical Score" = AGG_PHYS_2,
         "Baseline Log Viral Load" = logVLOAD_0,
         "Year 2 Log Viral Load" = logVLOAD_2,
         "Baseline CD4+ T Count" = LEU3N_0,
         "Year 2 CD4+ T Count" = LEU3N_2,
         "Baseline BMI" = BMI_0,
         "Baseline Education" = EDUCBAS_0,
         "Smoking Status at Baseline" = SMOKE_0,
         "Age at Baseline" = age_0,
         "Adherence to Treatment" = ADH_2,
         "Hard Drugs" = hard_drugs_0,
         "Race" = RACE_0)
  
dfTableOne %>% tbl_summary(by="Hard Drugs") %>%
    as_kable_extra(booktabs = TRUE) %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down"` argument here.
  kableExtra::kable_styling(font_size = 10)
```

Subject demographics are reported in table 1. Adherence was higher in hard drug users (95-100% Adherence: 96%) than in non hard drug users (95-100% Adherence: 89%). 

```{r, echo=FALSE}
dfBoxplot <- dfHIV %>% mutate(logVLOAD_0 = log(VLOAD_0, base=10),
                               logVLOAD_2 = log(VLOAD_2, base=10),
                               MENT_DIFF = AGG_MENT_2 - AGG_MENT_0,
                               PHYS_DIFF = AGG_PHYS_2 - AGG_PHYS_0,
                               logVLOAD_DIFF = logVLOAD_2 - logVLOAD_0,
                               LEU3N_DIFF = LEU3N_2 - LEU3N_0,) %>%
  select(hard_drugs_0, MENT_DIFF, PHYS_DIFF, logVLOAD_DIFF, LEU3N_DIFF) %>% 
  pivot_longer(cols=-c(hard_drugs_0), names_to="Outcome", values_to="Value") %>%
  mutate(Outcome = case_match(Outcome, 
                              "MENT_DIFF" ~ "Difference in Aggregate Mental Score (Points)",
                              "PHYS_DIFF" ~ "Difference in Aggregate Physical Score (Points)",
                              "LEU3N_DIFF" ~ "Difference in CD4+ T Cell Count (Cells)",
                              "logVLOAD_DIFF" ~ "Difference in log(Viral Load) (log(copies/mL))"))
dfBoxplot %>% ggplot(aes(group=hard_drugs_0, y=Value, fill=Outcome)) +
  geom_boxplot() +
  facet_wrap(~Outcome, scales = "free") +
  labs(y=NULL) +
  theme(legend.position="none")
```

## Effect of Hard Drug Use

The effect of hard drug use is summarized for both the frequentist and Bayesian approaches in table 2.

```{r, echo=FALSE}
dfResults <- read.csv("./StoredResults/ResultsTable.csv")
kable(dfResults, col.names = c("Variable", "Frequentist Estimate", "95% Confidence Interval", "p Value", "$\\Delta$ AIC", "Bayesian Estimate", "95% Credible Interval", "Posterior Probability", "$\\Delta$ LOO-IC"))
```

### Frequentist

Hard drug use was associated with a nonsignificant 0.29 point decrease in aggregate mental health score (95% CI: -3.75, 3.17, p = 0.870). Additionally, based on AIC, hard drug use does not improve model fit for mental health score by a meaningful amount ($\Delta$ AIC = +1.97). It was associated with a significant 3.34 point decrease in aggregate physical health score (95% CI: -6.06, -0.61, p = 0.017) and including hard drug use meaningfully improved model fit for physical health score as measured by AIC ($\Delta$ AIC = -3.85).  It was also associated with a significantly lower CD4+ T cell count, with 163.88	fewer cells (95% CI: -226.35, -101.41, p < 0.001) and including hard drug use improved model fit by a highly meaninful amount for T Cell count as measured by AIC ($\Delta$ AIC = -24.29). Finally, hard drug use was associated with a nonsignificant decrease of 0.06 log(viral load) (95% CI: (-0.46, 0.34), p = 0.770). Additionally, based on AIC, hard drug use does not improve model fit for log(viral load) by a meaningful amount ($\Delta$ AIC = +1.91).

### Bayesian 

Hard drug use was associated with a 0.29 point decrease in aggregate mental health score. There is a 95% probability that the effect of hard drug use is between a 3.75 point decrease and a 3.14 increase in aggregate mental health score. The posterior probability that hard drug use results in a clinically meaningful 2 point difference in score is 26.2%. Additionally, based on LOO-IC, hard drug use does not improve model fit for the mental health score by a meaningful amount ($\Delta$ LOO-IC = +3.00).

Hard drug use was associated with a 3.33 point decrease in aggregate physical health score. There is a 95% probability that the effect of hard drug use is between a 6.07 and 0.58 point decrease in aggregate physical health score. The posterior probability that hard drug use results in a clinically meaningful 2 point difference in score is 83.1%. Additionally, based on LOO-IC, hard drug use improved model fit for the physical health score by a meaningful amount ($\Delta$ LOO-IC = -3.45).

Hard drug use was associated with a 148.95 cell decrease in CD4+ T cell count. There is a 95% probability that the effect of hard drug use is between a 208.17 and 89.28 cell decrease in CD4+ T cell count. The posterior probability that hard drug use results in a clinically meaningful 50 cell difference is 99.9%. Additionally, based on LOO-IC, hard drug use improved model fit for T cell count by a highly meaningful amount ($\Delta$ LOO-IC = -23.49).

Hard drug use was associated with a 0.06 decrease in log(viral load). There is a 95% probability that the effect of hard drug use is between a 0.46 point decrease and a 0.34 increase in log(viral load). The posterior probability that hard drug use results in a clinically meaningful 0.5 difference in log(viral load) is 1.8%. Additionally, based on LOO-IC, hard drug use does not improve model fit for log(viral load) by a meaningful amount ($\Delta$ LOO-IC = +2.04).

## Mediation by Adherence

The impact of adherence on the effect of hard drug use is summarized for both the frequentist and Bayesian approaches in table 3.

```{r, echo=FALSE}
dfResultsAdherence <- read.csv("./StoredResults/ResultsAdherenceTable.csv")
kable(dfResultsAdherence, col.names = c("Variable", "Frequentist Estimate (95% CI)", "Frequentist Estimate with Adherence (95% CI)", "$\\Delta$ Estimate", "$\\Delta$ AIC", "Bayesian Estimate (95% CI)", "Bayesian Estimate with Adherence (95% CI)", "$\\Delta$ Estimate", "$\\Delta$ LOO-IC"))
```

### Frequentist

The change in aggregate mental health score was 0.24 points lower in the model with adherence. However, based on AIC, the inclusion of adherence did not meaningfully improve the model fit for aggregate mental health score as the 0.26 change in AIC was less than 3. The change in aggregate physical health score was 0.23 points lower in the model with adherence. However, based on AIC, the inclusion of adherence did not meanginfully improve the model fit for aggregate physcial health score as the 1.31 change in AIC was less than 3. The change in T cell count was 5.8 cells lower in the model with adherence. However, based on AIC, the inclusion of adherence did not meaningfully improve the model fit for T cell count as the 1.98 change in AIC was less than 3. The change in log(viral load) was 0.05 points higher in the model with adherence and, based on AIC, the inclusion of adherence meaningfully improved the model fit for log(viral load) with a 4.27 change in AIC.

### Bayesian

The change in aggregate mental health score was 0.25 points lower in the model with adherence. However, the inclusion of adherence did not improve the model as the 0.02 change in LOO-IC was less than 3. The change in aggregate physical health score was 0.24 points lower in the model with adherence. However, the inclusion of adherence did not improve the model as the 1.56 change in LOO-IC was less than 3. The change in T cell count was 4.83 cells lower in the model with adherence. However, the inclusion of adherence did not improve the model as the 2.30 change in LOO-IC was less than 3. The change in log(viral load) was 0.05 points higher in the model with adherence and the inclusion of adherence improved the model with a 3.88 change in LOO-IC.

<!-- Tables and figures can come at the end of the report or be interleaved. -->

<!-- Table 1 is a table describing the dataset and often with separate summaries by the primary variable of interest (e.g., treatment group, disease/no-disease, other as defined by hypothesis). This table usually doesn't have p-values, but sometimes for space limitations in journals, will have p-values. Be careful about whether you are using standard deviations vs. standard errors. SD is used without p-values. SE with p-values. -->

<!-- Tables 2-X are tables of results.  I usually have a presentation of basic findings prior to adjustment next to findings with adjustment. How the tables are organized will be project specific. -->

<!-- Graphic/plot ideas: histograms by variable of interest. Bar chars of major findings with means and 95% CI and p-values. Fitted regression lines for regression based analysis. Color is fine for class. Remember to ask an investigator if they need black and white for publication. -->

<!-- Total 6 tables and figures.  Outside of class, more than this would be an appendix or up for discussion on what to keep.  The number of tables and figures is limited in nearly all journals. -->

<!-- First paragraph are descriptive statistics and graphs with a description of the patterns of importance in a table/figures describing the data. -->

<!-- ## Associations between primary outcome and treatment -->

<!-- Go on to describe the results from your crude analyses and then your adjusted analyses.  This should include a reference to a results table, differences in the outcome by treatment group with estimates, confidence intervals and p-values. You might then compare how these crude results change with adjustment for precision/confounding variables (your final models). -->

<!-- ## Additional subheaders as necessary to guide through a more complex analysis. -->

# Discussion and Summary

In both frequentist and Bayesian analysis, hard drug use was found to be associated with lower aggregate physical health scores and CD4+ T cell counts. It was not associated with aggregate mental health scores or log viral load. The association with CD4+ T cell counts was especially strong, and supports the prior finding in other species that hard drugs may reduce immune functionality. Reduced immune functioning may also be related to why the physical scores were lower or this result may be related to other mechanisms, as causality cannot be established from this study and many factors determine physical health. However, it is unclear why viral load was not significantly higher in people with hard drug use given that the measure of immune system was significantly lower. Further biological insight on the mechanisms of HIV and its treatment is required to understand whether or not this finding would be plausible. 

The mediation analysis found that only log viral load was better explained when including adherence as a covariate, as measured by AIC and LOO-IC in the frequentist and Bayesian models respectively. However, hard drug use was not significantly associated with log viral load regardless. For all other outcomes, including adherence did not improve model fit. 

One major limitation of the study was that there was a large amount of drop out between year 0 and year 2. Dropout is frequently non-random and very likely to be related to having a worse outcome in a study of disease progression. If drop out rates differ between hard drug users and non-hard drug users, then the study is likely to be biased by this. A more complex analysis is needed to take this possibility into account. Additionally, race, smoking status, and education level were each collapsed to two categories, which may have limited their efficacy in controlling for variability and reduced the precision of the analysis. QQ plots also indicated some divergence in the tails of the models for the physical and mental health scores and log viral load. Data transformation or more complex models may need to be considered for these outcomes. The study is also unable to establish a causative relationship between hard drug use and the differences in outcome, as it is still possible that other unmeasured confounding factors caused both hard drug use and more negative outcomes. 

Adherence was also collapsed to only two categories, <95% and 95-100%, which may have limited the mediation analysis. Given that hard drug users had higher adherence, it is possible that their higher adherence could have masked the effect of using hard drugs if the trend did not follow the binary adherence categories. A more detailed future mediation analysis may be worthwhile.

# Reproducibility

Code is available at https://github.com/isabellemeredith/BIOS6624/tree/main/Project1
Data was located at "/Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project1/DataRaw/hiv_6624_final.csv"
