---
title: "Project 4 Report"
author: "Isabelle Meredith"
date: "2026-04-29"
header-includes: 
  \usepackage[figuresright]{rotating}
  \usepackage{lscape}
  \usepackage{pdfpages}
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(hdrm)
library(glmnet)
library(StepReg)
library(dplyr)
library(tidyr)
library(gtsummary)
library(gt)
library(kableExtra)
```


## Introduction

Variable selection is frequently used in research, both to look for potential relationships in an exploratory analysis and to avoid overfitting when controlling for covariates during hypothesis testing. Accordingly, many forms of variable selection have developed. However, these methods may vary in efficacy and have the potential to introduce bias. In order to assess the efficacy and bias of variable selection methods, we will simulate data with known relationships and evaluate how often the variables with nonzero relationships are selected as well as the bias of the estimates for the selected variables. The methods assessed will be backward selection based on F-tests, p-values, AIC, and BIC, along with lasso, and elastic net. 

### Research Question

How effective are the most common variable selection methods and how much bias do they introduce to estimates?

## Methods

### Data Simulation

Data was generated using the hdrm package for R. Two cases were considered, a case with 250 subjects and a case with 500 subjects. In both cases we generated 20 variables of which 5 had non-zero relationships with the outcome. The coefficients of these 5 variables were 1/6, 1/3, 1/2, 2/3, and 5/6 respectively. The remaining 15 variables had no relationship with the outcome. We generated a version of each case with an exchangeable correlation of 0, an exchangeable correlation of 0.35, and an exchangeable correlation of 0.7, for a total of 6 combinations. 10,000 datasets were created for each combination.

### Data Analysis

We tested five different variable selection methods: backward selection based on p-values, AIC, and BIC, along with lasso, and elastic net. For the backward selection methods with p-values a custom function was used. The threshold was 0.05 for p values. For AIC and BIC, we performed variable selection using the package StepReg. Any reduction in AIC or BIC was considered sufficient. For lasso and elastic net, models were fit with the package glmnet. Two approaches for selecting lambda were used with cross validation: the minimum MSE and the lowest lambda within 1 standard deviation of the minimum MSE. A value of 0.5 was used for alpha in elastic net. A final model was refit with the included variables. For each method, we counted the number of variables with non-zero relationship that were correctly included and the number of variables with no relationship that were incorrectly included. For each variable that was retained in the final model, we recorded the parameter estimate, 95% CI, and p value. This procedure was repeated for each case of the datasets specified in data simulation. Using the collective results for each dataset case and method, we calculated the bias of the parameter estimates, the true coverage of the 95% CI, and the type I and II error rates.

```{r, echo=FALSE, eval=FALSE}
variables <- colnames(gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)))$X)
methods <- c("PValue", "AIC", "BIC", "LassoMin", "Lasso1SD", "ElasticMin", "Elastic1SD")
values <- c("Coef", "p", "CIL", "CIU")

# start with all variables
# get max p
# if max p is greater than the minimum threshold, remove it
# if not, end loop
# return model
pValueSelection <- function(variables, data, cutoff=0.05) {
  maxPValue <- 1
  includedVariables <- variables
  while (maxPValue > cutoff) {
    modelformula = as.formula(paste("y ~", paste(includedVariables, collapse = " + "), "+ -1"))
    model <- lm(modelformula, data)
    ps <- summary(model)$coefficients[,4]
    maxPValue <- max(ps)
    maxPValueVariable <- row.names(summary(model)$coefficients)[which.max(ps)]
    if (maxPValue > cutoff) {
      includedVariables <- includedVariables[includedVariables != maxPValueVariable]
    }
  }
  return(model)
}

refitLM <- function(coef, data) {
  coefnames <- variables[which(coef != 0)-1]
  formula <- as.formula(paste("y ~", paste(coefnames, collapse = " + "), "+ -1"))
  lm(formula, data=data)
}

selectionLoopInterior <- function(inputData) {
  dfCoefficientsPValue <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsAIC <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsBIC <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsLassoMin <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsLasso1SD <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsElasticMin <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsElastic1SD <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  
  colnames(dfCoefficientsPValue) <- paste(values, rep(paste("PValue", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsAIC) <- paste(values, rep(paste("AIC", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsBIC) <- paste(values, rep(paste("BIC", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsLassoMin) <- paste(values, rep(paste("LassoMin", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsLasso1SD) <- paste(values, rep(paste("Lasso1SD", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsElasticMin) <- paste(values, rep(paste("ElasticMin", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsElastic1SD) <- paste(values, rep(paste("Elastic1SD", variables, sep="-"), each=4), sep="-")
  
  data <- data.frame(inputData$X)
  data$y <- inputData$y
  
  modelPValue <- pValueSelection(variables, data)
  
  selectionAIC <- stepwise(y ~ . + -1, data=data, strategy="backward", metric="AIC")
  selectionBIC <- stepwise(y ~ . + -1, data=data, strategy="backward", metric="BIC")
  
  modelAIC <- lm(selectionAIC$performance$model, data=data)
  modelBIC <- lm(selectionBIC$performance$model, data=data)

  lasso <- cv.glmnet(inputData$X, inputData$y, alpha=1, intercept=FALSE)

  lassoMin <- refitLM(coef(lasso, s= 'lambda.min'), data)
  lasso1SD <- refitLM(coef(lasso, s= 'lambda.1se'), data)
  
  elasticnet <- cv.glmnet(inputData$X, inputData$y, alpha=0.5, intercept=FALSE)

  elasticnetMin <- refitLM(coef(elasticnet, s= 'lambda.min'), data)
  elasticnet1SD <- refitLM(coef(elasticnet, s= 'lambda.1se'), data)
  
  dfCoefficientsPValue[1, paste("Coef-PValue", names(modelPValue$coef), sep="-")] <- unname(modelPValue$coef)
  dfCoefficientsAIC[1, paste("Coef-AIC", names(modelAIC$coef), sep="-")] <- unname(modelAIC$coef)
  dfCoefficientsBIC[1, paste("Coef-BIC", names(modelBIC$coef), sep="-")] <- unname(modelBIC$coef)
  dfCoefficientsLassoMin[1, paste("Coef-LassoMin", names(lassoMin$coef), sep="-")] <- unname(lassoMin$coef)
  dfCoefficientsLasso1SD[1, paste("Coef-Lasso1SD", names(lasso1SD$coef), sep="-")] <- unname(lasso1SD$coef)
  dfCoefficientsElasticMin[1, paste("Coef-ElasticMin", names(elasticnetMin$coef), sep="-")] <- unname(elasticnetMin$coef)
  dfCoefficientsElastic1SD[1, paste("Coef-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- unname(elasticnet1SD$coef)
  
  dfCoefficientsPValue[1, paste("p-PValue", names(modelPValue$coef), sep="-")] <- summary(modelPValue)$coef[,4] 
  dfCoefficientsAIC[1, paste("p-AIC", names(modelAIC$coef), sep="-")] <- summary(modelAIC)$coef[,4] 
  dfCoefficientsBIC[1, paste("p-BIC", names(modelBIC$coef), sep="-")] <- summary(modelBIC)$coef[,4] 
  dfCoefficientsLassoMin[1, paste("p-LassoMin", names(lassoMin$coef), sep="-")] <- summary(lassoMin)$coef[,4] 
  dfCoefficientsLasso1SD[1, paste("p-Lasso1SD", names(lasso1SD$coef), sep="-")] <- summary(lasso1SD)$coef[,4] 
  dfCoefficientsElasticMin[1, paste("p-ElasticMin", names(elasticnetMin$coef), sep="-")] <- summary(elasticnetMin)$coef[,4] 
  dfCoefficientsElastic1SD[1, paste("p-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- summary(elasticnet1SD)$coef[,4] 
  
  dfCoefficientsPValue[1, paste("CIL-PValue", names(modelPValue$coef), sep="-")] <- confint(modelPValue)[,1]
  dfCoefficientsAIC[1, paste("CIL-AIC", names(modelAIC$coef), sep="-")] <- confint(modelAIC)[,1]
  dfCoefficientsBIC[1, paste("CIL-BIC", names(modelBIC$coef), sep="-")] <- confint(modelBIC)[,1]
  dfCoefficientsLassoMin[1, paste("CIL-LassoMin", names(lassoMin$coef), sep="-")] <- confint(lassoMin)[,1]
  dfCoefficientsLasso1SD[1, paste("CIL-Lasso1SD", names(lasso1SD$coef), sep="-")] <- confint(lasso1SD)[,1]
  dfCoefficientsElasticMin[1, paste("CIL-ElasticMin", names(elasticnetMin$coef), sep="-")] <- confint(elasticnetMin)[,1]
  dfCoefficientsElastic1SD[1, paste("CIL-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- confint(elasticnet1SD)[,1]
  
  dfCoefficientsPValue[1, paste("CIU-PValue", names(modelPValue$coef), sep="-")] <- confint(modelPValue)[,2]
  dfCoefficientsAIC[1, paste("CIU-AIC", names(modelAIC$coef), sep="-")] <- confint(modelAIC)[,2]
  dfCoefficientsBIC[1, paste("CIU-BIC", names(modelBIC$coef), sep="-")] <- confint(modelBIC)[,2]
  dfCoefficientsLassoMin[1, paste("CIU-LassoMin", names(lassoMin$coef), sep="-")] <- confint(lassoMin)[,2]
  dfCoefficientsLasso1SD[1, paste("CIU-Lasso1SD", names(lasso1SD$coef), sep="-")] <- confint(lasso1SD)[,2]
  dfCoefficientsElasticMin[1, paste("CIU-ElasticMin", names(elasticnetMin$coef), sep="-")] <- confint(elasticnetMin)[,2]
  dfCoefficientsElastic1SD[1, paste("CIU-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- confint(elasticnet1SD)[,2]
  
  return(c(dfCoefficientsPValue, dfCoefficientsAIC, dfCoefficientsBIC, dfCoefficientsLassoMin, dfCoefficientsLasso1SD, dfCoefficientsElasticMin, dfCoefficientsElastic1SD))
}

```

```{r echo=FALSE, eval=FALSE}
set.seed(618)
nIter <- 10000

dfcase1a <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase1a) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")

for (i in 1:nIter) {
  case1a <- gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)))
  
  dfList <- selectionLoopInterior(case1a)
  dfcase1a[i,] <- dfList
}
write.csv(dfcase1a, "./Data/case1a.csv")
rm(dfcase1a)

dfcase1b <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase1b) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case1b <- gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.35)
  dfList <- selectionLoopInterior(case1b)
  dfcase1b[i,] <- dfList
}
write.csv(dfcase1b, "./Data/case1b.csv")
rm(dfcase1b)
  
dfcase1c <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase1c) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case1c <- gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.7)
  dfList <- selectionLoopInterior(case1c)
  dfcase1c[i,] <- dfList
}
write.csv(dfcase1c, "./Data/case1c.csv")
rm(dfcase1c)
  
dfcase2a <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase2a) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case2a <- gen_data(500, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)))
  dfList <- selectionLoopInterior(case2a)
  dfcase2a[i,] <- dfList
}
write.csv(dfcase2a, "./Data/case2a.csv")
rm(dfcase2a)
  
dfcase2b <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase2b) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case2b <- gen_data(500, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.35)
  dfList <- selectionLoopInterior(case2b)
  dfcase2b[i,] <- dfList
}
write.csv(dfcase2b, "./Data/case2b.csv")
rm(dfcase2b)  
  
dfcase2c <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase2c) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case2c <- gen_data(500, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.7)
  dfList <- selectionLoopInterior(case2c)
  dfcase2c[i,] <- dfList
}
write.csv(dfcase2c, "./Data/case2c.csv")
```

## Results

The bias and coverage of each method on each dataset is given by variable in table 1. The frequency of inclusion and the power/type I error for each variable for each method on each dataset is given in table 2.

```{r echo=FALSE, warning=FALSE, message=FALSE}
cols_order <- c("Method", "Dataset", unlist(lapply(c("V01", "V02", "V03", "V04", "V05", "Other"), function(x) paste(x, c("Bias", "Cover", "Incl", "Pow"), sep = "_"))))

summary <- data.frame(matrix(nrow=0,ncol=length(cols_order),dimnames=list(NULL,cols_order))) %>% 
  mutate(Method = as.character(Method),
         Dataset = as.character(Dataset))

datasets <- c("case1a", "case1b", "case1c", "case2a", "case2b", "case2c")

for (dataset in datasets) {
  data <- read.csv(paste("./Data/", dataset, ".csv", sep=""))
  dataSummary <- data %>% pivot_longer(everything(), names_pattern="(.*)\\.(V.*)", names_to = c(".value", "Variable")) %>%
  pivot_longer(-c(Variable), names_sep="\\.", names_to = c(".value", "Method")) %>%
  mutate(Variable = ifelse(Variable %in% c("V01", "V02", "V03", "V04", "V05"), Variable, "Other"),
         covered = case_match(Variable,
                                 "V01" ~ (CIL < 1/6) & (CIU > 1/6),
                                 "V02" ~ (CIL < 1/3) & (CIU > 1/3),
                                 "V03" ~ (CIL < 1/2) & (CIU > 1/2),
                                 "V04" ~ (CIL < 2/3) & (CIU > 2/3),
                                 "V05" ~ (CIL < 5/6) & (CIU > 5/6),
                                 .default = (CIL < 0) & (CIU > 0))) %>%
  group_by(Method, Variable) %>%
  summarise(n = n(),
            Coef = mean(Coef, na.rm=TRUE),
            Pow = sum((p < 0.05) & !is.na(p)) / n * 100,
            Incl = sum(!is.na(p)) / n * 100,
            Cover = sum(covered & !is.na(covered)) / n * 100) %>% 
  mutate(Bias = case_match(Variable,
                                 "V01" ~ Coef - 1/6,
                                 "V02" ~ Coef - 1/3,
                                 "V03" ~ Coef - 1/2,
                                 "V04" ~ Coef - 2/3,
                                 "V05" ~ Coef - 5/6,
                                 .default = Coef),
         Dataset = dataset) %>%
  select(-c(n, Coef)) %>%
  pivot_wider(names_from = "Variable", values_from = c("Bias", "Cover", "Incl", "Pow"), names_glue = "{Variable}_{.value}") %>%
  select(all_of(cols_order))
  
  summary <- bind_rows(summary, dataSummary)
}
```

```{r echo=FALSE}
summary %>% select(-contains("Incl"), -contains("Pow")) %>% 
  gt(rowname_col = "Method", groupname_col = "Dataset") %>%
  tab_spanner_delim(
    delim = "_"
  ) %>%
  tab_options(table.font.size = 12) %>%
  fmt_number(decimals=2)
```

```{r echo=FALSE}
summary %>% select(-contains("Bias"), -contains("Cover")) %>%
  gt(rowname_col = "Method", groupname_col = "Dataset") %>%
  tab_spanner_delim(
    delim = "_"
  ) %>%
  tab_options(table.font.size = 12) %>%
  fmt_number(decimals=2)
```

## Discussion

what portion of the true variables were included? for the rates

you should definitely report on unconditional bias. conditional bias is also important, but I will leave it up to you as to whether you want to discuss it

## Reproducibility

Github Repository: https://github.com/isabellemeredith/BIOS6624

Data Directory: /Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project4/Data
