---
title: "Project 4 Report"
author: "Isabelle Meredith"
date: "2026-04-29"
header-includes: 
  \usepackage[figuresright]{rotating}
  \usepackage{lscape}
  \usepackage{pdfpages}
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(hdrm)
library(glmnet)
library(StepReg)
library(dplyr)
library(tidyr)
library(gtsummary)
library(gt)
library(kableExtra)
```


## Introduction

Variable selection is frequently used in research, both to look for potential relationships in an exploratory analysis and to avoid overfitting when controlling for covariates during hypothesis testing. Accordingly, many forms of variable selection have developed. However, these methods may vary in efficacy and have the potential to introduce bias. In order to assess the efficacy and bias of variable selection methods, we will simulate data with known relationships and evaluate how often the variables with nonzero relationships are selected as well as the bias of the estimates for the selected variables. The methods assessed will be backward selection based on F-tests, p-values, Akaike Information Criterion (AIC), and Bayesian Information Criterion (BIC), along with lasso, and elastic net. 

### Research Question

How do common variable selection methods affect bias, coverage, and error rates?

## Methods

### Model Notation

$$Y = \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5 + \beta_6X_6 ... + \beta_{20}X_{20} + \epsilon$$
### Variable Selection Methods

We tested five different variable selection methods: backward selection based on p-values, AIC, and BIC, along with lasso, and elastic net. Variable selection based on p-values removes variables if they have a p value below a predetermined threshold. Selection based on AIC or BIC   

### Data Simulation

Data was generated using the hdrm package for R. Two cases were considered, a case with 250 subjects and a case with 500 subjects. In both cases we generated 20 variables of which 5 had non-zero relationships with the outcome. The coefficients of these 5 variables were 1/6, 1/3, 1/2, 2/3, and 5/6 respectively. The remaining 15 variables had no relationship with the outcome. We generated a version of each case with an exchangeable correlation of 0, an exchangeable correlation of 0.35, and an exchangeable correlation of 0.7, for a total of 6 combinations. 10,000 datasets were created for each combination.

For the backward selection methods with p-values a custom function was used. The threshold was 0.05 for p values. For AIC and BIC, we performed variable selection using the package StepReg. Any reduction in AIC or BIC was considered sufficient. For lasso and elastic net, models were fit with the package glmnet. Two approaches for selecting lambda were used with cross validation: the minimum MSE and the lowest lambda within 1 standard deviation of the minimum MSE. A value of 0.5 was used for alpha in elastic net. A final model was refit with the included variables. For each method, we counted the number of variables with non-zero relationship that were correctly included and the number of variables with no relationship that were incorrectly included. For each variable that was retained in the final model, we recorded the parameter estimate, 95% CI, and p value. This procedure was repeated for each case of the datasets specified in data simulation. Using the collective results for each dataset case and method, we calculated the bias of the parameter estimates, the true coverage of the 95% CI, and the type I and II error rates. These measures were calculated only when the variables were included, as it was not possible to get estimates for unselected variables for all models and the results needed to be comparable across methods. Additionally, this study is investigating the results of a naive selection.

```{r, echo=FALSE, eval=FALSE}
variables <- colnames(gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)))$X)
methods <- c("PValue", "AIC", "BIC", "LassoMin", "Lasso1SD", "ElasticMin", "Elastic1SD")
values <- c("Coef", "p", "CIL", "CIU")

# start with all variables
# get max p
# if max p is greater than the minimum threshold, remove it
# if not, end loop
# return model
pValueSelection <- function(variables, data, cutoff=0.05) {
  maxPValue <- 1
  includedVariables <- variables
  while (maxPValue > cutoff) {
    modelformula = as.formula(paste("y ~", paste(includedVariables, collapse = " + "), "+ -1"))
    model <- lm(modelformula, data)
    ps <- summary(model)$coefficients[,4]
    maxPValue <- max(ps)
    maxPValueVariable <- row.names(summary(model)$coefficients)[which.max(ps)]
    if (maxPValue > cutoff) {
      includedVariables <- includedVariables[includedVariables != maxPValueVariable]
    }
  }
  return(model)
}

refitLM <- function(coef, data) {
  coefnames <- variables[which(coef != 0)-1]
  formula <- as.formula(paste("y ~", paste(coefnames, collapse = " + "), "+ -1"))
  lm(formula, data=data)
}

selectionLoopInterior <- function(inputData) {
  dfCoefficientsPValue <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsAIC <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsBIC <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsLassoMin <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsLasso1SD <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsElasticMin <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  dfCoefficientsElastic1SD <- data.frame(matrix(data=rep(NA, 20*4), nrow=1, ncol=20*4))
  
  colnames(dfCoefficientsPValue) <- paste(values, rep(paste("PValue", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsAIC) <- paste(values, rep(paste("AIC", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsBIC) <- paste(values, rep(paste("BIC", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsLassoMin) <- paste(values, rep(paste("LassoMin", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsLasso1SD) <- paste(values, rep(paste("Lasso1SD", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsElasticMin) <- paste(values, rep(paste("ElasticMin", variables, sep="-"), each=4), sep="-")
  colnames(dfCoefficientsElastic1SD) <- paste(values, rep(paste("Elastic1SD", variables, sep="-"), each=4), sep="-")
  
  data <- data.frame(inputData$X)
  data$y <- inputData$y
  
  modelPValue <- pValueSelection(variables, data)
  
  selectionAIC <- stepwise(y ~ . + -1, data=data, strategy="backward", metric="AIC")
  selectionBIC <- stepwise(y ~ . + -1, data=data, strategy="backward", metric="BIC")
  
  modelAIC <- lm(selectionAIC$performance$model, data=data)
  modelBIC <- lm(selectionBIC$performance$model, data=data)

  lasso <- cv.glmnet(inputData$X, inputData$y, alpha=1, intercept=FALSE)

  lassoMin <- refitLM(coef(lasso, s= 'lambda.min'), data)
  lasso1SD <- refitLM(coef(lasso, s= 'lambda.1se'), data)
  
  elasticnet <- cv.glmnet(inputData$X, inputData$y, alpha=0.5, intercept=FALSE)

  elasticnetMin <- refitLM(coef(elasticnet, s= 'lambda.min'), data)
  elasticnet1SD <- refitLM(coef(elasticnet, s= 'lambda.1se'), data)
  
  dfCoefficientsPValue[1, paste("Coef-PValue", names(modelPValue$coef), sep="-")] <- unname(modelPValue$coef)
  dfCoefficientsAIC[1, paste("Coef-AIC", names(modelAIC$coef), sep="-")] <- unname(modelAIC$coef)
  dfCoefficientsBIC[1, paste("Coef-BIC", names(modelBIC$coef), sep="-")] <- unname(modelBIC$coef)
  dfCoefficientsLassoMin[1, paste("Coef-LassoMin", names(lassoMin$coef), sep="-")] <- unname(lassoMin$coef)
  dfCoefficientsLasso1SD[1, paste("Coef-Lasso1SD", names(lasso1SD$coef), sep="-")] <- unname(lasso1SD$coef)
  dfCoefficientsElasticMin[1, paste("Coef-ElasticMin", names(elasticnetMin$coef), sep="-")] <- unname(elasticnetMin$coef)
  dfCoefficientsElastic1SD[1, paste("Coef-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- unname(elasticnet1SD$coef)
  
  dfCoefficientsPValue[1, paste("p-PValue", names(modelPValue$coef), sep="-")] <- summary(modelPValue)$coef[,4] 
  dfCoefficientsAIC[1, paste("p-AIC", names(modelAIC$coef), sep="-")] <- summary(modelAIC)$coef[,4] 
  dfCoefficientsBIC[1, paste("p-BIC", names(modelBIC$coef), sep="-")] <- summary(modelBIC)$coef[,4] 
  dfCoefficientsLassoMin[1, paste("p-LassoMin", names(lassoMin$coef), sep="-")] <- summary(lassoMin)$coef[,4] 
  dfCoefficientsLasso1SD[1, paste("p-Lasso1SD", names(lasso1SD$coef), sep="-")] <- summary(lasso1SD)$coef[,4] 
  dfCoefficientsElasticMin[1, paste("p-ElasticMin", names(elasticnetMin$coef), sep="-")] <- summary(elasticnetMin)$coef[,4] 
  dfCoefficientsElastic1SD[1, paste("p-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- summary(elasticnet1SD)$coef[,4] 
  
  dfCoefficientsPValue[1, paste("CIL-PValue", names(modelPValue$coef), sep="-")] <- confint(modelPValue)[,1]
  dfCoefficientsAIC[1, paste("CIL-AIC", names(modelAIC$coef), sep="-")] <- confint(modelAIC)[,1]
  dfCoefficientsBIC[1, paste("CIL-BIC", names(modelBIC$coef), sep="-")] <- confint(modelBIC)[,1]
  dfCoefficientsLassoMin[1, paste("CIL-LassoMin", names(lassoMin$coef), sep="-")] <- confint(lassoMin)[,1]
  dfCoefficientsLasso1SD[1, paste("CIL-Lasso1SD", names(lasso1SD$coef), sep="-")] <- confint(lasso1SD)[,1]
  dfCoefficientsElasticMin[1, paste("CIL-ElasticMin", names(elasticnetMin$coef), sep="-")] <- confint(elasticnetMin)[,1]
  dfCoefficientsElastic1SD[1, paste("CIL-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- confint(elasticnet1SD)[,1]
  
  dfCoefficientsPValue[1, paste("CIU-PValue", names(modelPValue$coef), sep="-")] <- confint(modelPValue)[,2]
  dfCoefficientsAIC[1, paste("CIU-AIC", names(modelAIC$coef), sep="-")] <- confint(modelAIC)[,2]
  dfCoefficientsBIC[1, paste("CIU-BIC", names(modelBIC$coef), sep="-")] <- confint(modelBIC)[,2]
  dfCoefficientsLassoMin[1, paste("CIU-LassoMin", names(lassoMin$coef), sep="-")] <- confint(lassoMin)[,2]
  dfCoefficientsLasso1SD[1, paste("CIU-Lasso1SD", names(lasso1SD$coef), sep="-")] <- confint(lasso1SD)[,2]
  dfCoefficientsElasticMin[1, paste("CIU-ElasticMin", names(elasticnetMin$coef), sep="-")] <- confint(elasticnetMin)[,2]
  dfCoefficientsElastic1SD[1, paste("CIU-Elastic1SD", names(elasticnet1SD$coef), sep="-")] <- confint(elasticnet1SD)[,2]
  
  return(c(dfCoefficientsPValue, dfCoefficientsAIC, dfCoefficientsBIC, dfCoefficientsLassoMin, dfCoefficientsLasso1SD, dfCoefficientsElasticMin, dfCoefficientsElastic1SD))
}

```

```{r echo=FALSE, eval=FALSE}
set.seed(618)
nIter <- 10000

dfcase1a <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase1a) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")

for (i in 1:nIter) {
  case1a <- gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)))
  
  dfList <- selectionLoopInterior(case1a)
  dfcase1a[i,] <- dfList
}
write.csv(dfcase1a, "./Data/case1a.csv")
rm(dfcase1a)

dfcase1b <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase1b) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case1b <- gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.35)
  dfList <- selectionLoopInterior(case1b)
  dfcase1b[i,] <- dfList
}
write.csv(dfcase1b, "./Data/case1b.csv")
rm(dfcase1b)
  
dfcase1c <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase1c) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case1c <- gen_data(250, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.7)
  dfList <- selectionLoopInterior(case1c)
  dfcase1c[i,] <- dfList
}
write.csv(dfcase1c, "./Data/case1c.csv")
rm(dfcase1c)
  
dfcase2a <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase2a) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case2a <- gen_data(500, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)))
  dfList <- selectionLoopInterior(case2a)
  dfcase2a[i,] <- dfList
}
write.csv(dfcase2a, "./Data/case2a.csv")
rm(dfcase2a)
  
dfcase2b <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase2b) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case2b <- gen_data(500, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.35)
  dfList <- selectionLoopInterior(case2b)
  dfcase2b[i,] <- dfList
}
write.csv(dfcase2b, "./Data/case2b.csv")
rm(dfcase2b)  
  
dfcase2c <- data.frame(matrix(data=rep(NA, 20*7*4*10000), nrow=10000, ncol=20*4*7))
colnames(dfcase2c) <- paste(values, rep(paste(rep(methods, each = length(variables)), variables, sep = "-"), each=4), sep = "-")
for (i in 1:nIter) {
  case2c <- gen_data(500, 20, 5, beta=append(c(1/6, 1/3, 1/2, 2/3, 5/6), rep(0,15)), corr="exchangeable", rho=0.7)
  dfList <- selectionLoopInterior(case2c)
  dfcase2c[i,] <- dfList
}
write.csv(dfcase2c, "./Data/case2c.csv")
```

## Results

The bias and coverage of each method on each dataset is given by variable in table 1. For case 1a and 2a with no correlation among variables, there was little to no bias for all variables. V01, which had the smallest coefficient of 1/6, was the only variable which had any bias. Bias was between 0.01 and 0.04, particularly on the 1SD lambda choice for lasso and elastic net. For the remaining datasets, some bias was present for at least one variable for all methods. The most bias was present on the lasso and elastic net methods, regardless of the method of choosing lambda, except in 1c and potentially 2c on V01 where AIC and BIC had high bias as well. 

The frequency of inclusion and the power/type I error for each variable for each method on each dataset is given in table 2. The power is extremely high for V03-V05 across all methods and datasets. V02 had a lower but reasonable power between 67% and 100% for all datasets. V01 had a reasonable power in dataset 1a, 2a, and 2b for all methods between 66% and 95%, except in lasso with 1SD lambda choice, where it dropped to 57%. The power for V01 was low for the remaining datasets. The lowest power was found in the lasso and elasticnet selection methods, particularly the 1SD lambda choice for both methods. However, the lasso and elasticnet methods with 1SD lambda choice also had the lowest type I error across all of the datasets. Selection with AIC and BIC had the highest type I error, which was often above the expected 5% and was worse in the smaller datasets and when correlation was higher. Selection with a p value of 0.05 was the closest to the expected 5% type I error rate.

```{r echo=FALSE, warning=FALSE, message=FALSE}
cols_order <- c("Method", "Dataset", unlist(lapply(c("V01", "V02", "V03", "V04", "V05", "Other"), function(x) paste(x, c("Bias", "Cover", "Incl", "Pow"), sep = "_"))))

summary <- data.frame(matrix(nrow=0,ncol=length(cols_order),dimnames=list(NULL,cols_order))) %>% 
  mutate(Method = as.character(Method),
         Dataset = as.character(Dataset))

datasets <- c("case1a", "case1b", "case1c", "case2a", "case2b", "case2c")

for (dataset in datasets) {
  data <- read.csv(paste("./Data/", dataset, ".csv", sep=""))
  dataSummary <- data %>% pivot_longer(everything(), names_pattern="(.*)\\.(V.*)", names_to = c(".value", "Variable")) %>%
  pivot_longer(-c(Variable), names_sep="\\.", names_to = c(".value", "Method")) %>%
  mutate(Variable = ifelse(Variable %in% c("V01", "V02", "V03", "V04", "V05"), Variable, "Other"),
         covered = case_match(Variable,
                                 "V01" ~ (CIL < 1/6) & (CIU > 1/6),
                                 "V02" ~ (CIL < 1/3) & (CIU > 1/3),
                                 "V03" ~ (CIL < 1/2) & (CIU > 1/2),
                                 "V04" ~ (CIL < 2/3) & (CIU > 2/3),
                                 "V05" ~ (CIL < 5/6) & (CIU > 5/6),
                                 .default = (CIL < 0) & (CIU > 0))) %>%
  group_by(Method, Variable) %>%
  summarise(n = n(),
            Coef = mean(Coef, na.rm=TRUE),
            Pow = sum((p < 0.05) & !is.na(p)) / n * 100,
            Incl = sum(!is.na(p)) / n * 100,
            Cover = sum(covered & !is.na(covered)) / n * 100) %>% 
  mutate(Bias = case_match(Variable,
                                 "V01" ~ Coef - 1/6,
                                 "V02" ~ Coef - 1/3,
                                 "V03" ~ Coef - 1/2,
                                 "V04" ~ Coef - 2/3,
                                 "V05" ~ Coef - 5/6,
                                 .default = Coef),
         Dataset = dataset) %>%
  select(-c(n, Coef)) %>%
  pivot_wider(names_from = "Variable", values_from = c("Bias", "Cover", "Incl", "Pow"), names_glue = "{Variable}_{.value}") %>%
  select(all_of(cols_order))
  
  summary <- bind_rows(summary, dataSummary)
}
```

```{r echo=FALSE}
summary %>% select(-contains("Incl"), -contains("Pow")) %>% 
  gt(rowname_col = "Method", groupname_col = "Dataset") %>%
  tab_spanner_delim(
    delim = "_"
  ) %>%
  tab_options(table.font.size = 12) %>%
  fmt_number(decimals=2)
```

```{r echo=FALSE}
summary %>% select(-contains("Bias"), -contains("Cover")) %>%
  rename(Other_T1E = Other_Pow) %>%
  gt(rowname_col = "Method", groupname_col = "Dataset") %>%
  tab_spanner_delim(
    delim = "_"
  ) %>%
  tab_options(table.font.size = 12) %>%
  fmt_number(decimals=2)
```

## Discussion

The patterns of bias are likely due to the correlation that was induced between variables. The higher bias for V01 but not the other variables for the b and c cases for AIC and BIC selection is likely because it had the weakest effect and was therefore the most likely to be removed due to the correlation structure. 

<!-- what portion of the true variables were included? for the rates -->

<!-- you should definitely report on unconditional bias. conditional bias is also important, but I will leave it up to you as to whether you want to discuss it -->

### Future Work

Exploring correlation structures besides exchangeable would be valuable for future research, as real data often has more complex correlation structures. Exploring the effects in larger and smaller datasets would also be valuable, especially smaller datasets. Variable selection is frequently used when there is high data dimensionality and Lasso and Elasticnet were designed for cases when there are more features than individuals. 

## Reproducibility

Github Repository: https://github.com/isabellemeredith/BIOS6624

Data Directory: /Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project4/Data
