---
title: "Project 3 Report"
author: "Isabelle Meredith"
date: "2026-04-08"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(naniar)
library(dplyr)
library(tidyr)
library(kableExtra)
library(gtsummary)
library(ggplot2)
library(survival)
library(easysurv)
library(survminer)
```

## Introduction

Identifying risk factors for strokes is an important part of preventative care. In this exploratory study, we will search for potential predictors of stroke risk in the Framingham Heart Study cohort. The subset of the study data that we are investigating runs from 1956 to 1968 and contains 4,434 participants. Participants received examinations every six years, for a total of three examinations during the study time frame. Laboratory, clinic, and questionnaire data were collected, as well as the time to event if a stroke occurred. We will examine the risk of stroke within the ten year period after the start of the study. As risk factors may differ between males and females, the analysis will be stratified by sex. Additionally, we will examine whether the values of the risk factors we found at the baseline examination vary over the course of the study or if they remain constant. 

### Question 1
What factors predict risk of stroke over a 10 year period in males and females?

### Question 2
What is the probability of stroke over a 10 year period for different risk profiles?

### Question 3
Do an individual's risk factors change over time or do they remain similar for the duration of the study?

## Methods

### Data Cleaning and Processing

Participants with prior strokes were excluded from the analysis (n = 32) leaving 4402 participants. Participants who had strokes more than 10 years after the start of data collection were considered to be censored at the 10 year mark for the sake of the analysis, as were all participants who did not have a stroke. Assessment of data distribution and outliers was conducted using histograms and descriptive statistics. Some very high values for cholesterol (>600) and for systolic blood pressure (>250) were seen and were considered suspicious but as there were several distinct high values and these values are theoretically possible, they were not discarded. Additionally, as we are performing complete case analysis, participants with missing baseline values for blood pressure medication use (n = 60), total cholesterol (n = 51), or BMI (n = 16) were removed, leaving a total of 4275 participants.

### Data Analysis

We explored potential risk factors for strokes using Cox proportional hazard models. We modeled males and females separately as there may be differences in risk factor by sex. We included age, diabetes, and blood pressure as known risk factors. We also considered coronary heart disease status, blood pressure medication use, smoking status, total cholesterol, and BMI as potential risk factors for stroke. Suitability for proportional hazard ratio modeling were confirmed visually using log minus log curves and Schoenfeld residuals. If the effect of a variable was found to vary with time, an interaction with time was included in the model to account for the violated assumption. We performed backwards variable selection using AIC to determine which variables are associated with risk of stroke. A change in AIC of 2 or more was considered meaningful for variable selection. We performed selection separately for the models for males and females. 

The probability of having a stroke within 10 years was calculated using the finalized risk factor models for several risk profiles. The profiles included males and females with average or normal values for all risk factors at baseline ages of 40, 50, and 60. Additionally, stroke probability was calculated for males and females with a high systolic blood pressure value of 160, with diabetes, and with both high blood pressure and diabetes at baseline age 40, 50, and 60. Finally, a risk profile was calculated for individuals with previous coronary heart disease, based on the results of the study.

In order to explore potential change in risk factors, the change in the risk factors was quantified between each of the three time points. For continuous covariates, the median absolute difference between the time points was calculated to account for the possibility of change in either direction, and for categorical covariates, the number and percentage of participants who changed category was calculated.

## Results

Patient demographics are given in table 1. 

```{r echo=FALSE, results='hide'}
dfCardio <- read.csv("/Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project3/DataRaw/frmgham2.csv")

unique(dfCardio %>% filter(TIMESTRK == 0) %>% select(RANDID)) 

dfCardioFiltered <- dfCardio %>% filter(TIME == 0) %>%
  filter(TIMESTRK != 0) %>% 
  mutate(TIMESTRKORIGINAL = TIMESTRK,
         TIMESTRK = ifelse(TIMESTRK < 3653, TIMESTRK, 3653),
         STROKEORIGINAL = STROKE,
         STROKE = ifelse(((TIMESTRK < 3653) & (STROKE == 1)), 1, 0)) %>%
  # select(-c(HDLC, LDLC, TIMEAP, TIMEMI, TIMEMIFC, TIMECHD, TIMECVD, TIMEHYP)) %>%
  select(TIMESTRKORIGINAL, TIMESTRK, STROKEORIGINAL, STROKE, SEX, AGE, DIABETES, SYSBP, DIABP, PREVCHD, BPMEDS, CURSMOKE, TOTCHOL, BMI) %>% 
  filter(!is.na(BPMEDS)) %>% 
  filter(!is.na(TOTCHOL)) %>%
  filter(!is.na(BMI))

dfCardioMale <- dfCardioFiltered %>% filter(SEX == 1)
  
dfCardioFemale <- dfCardioFiltered %>% filter(SEX == 2)

length(unique(dfCardioFiltered$RANDID))
```

```{r echo=FALSE, fig.cap="Summary of patient demographics by sex.", fig.pos = "H", out.extra = ''}
dfTableOne <- dfCardio %>% filter(TIME == 0) %>%
  filter(TIMESTRK != 0) %>%
  select(SEX, AGE, TOTCHOL, SYSBP, DIABP, CURSMOKE, BMI, DIABETES, BPMEDS, PREVCHD, TIMESTRK, STROKE) %>%
  mutate(SEX = ifelse(SEX == 1, "Male", "Female"),
         CURSMOKE = case_match(CURSMOKE, 
                              0 ~ "No",
                              1 ~ "Yes"),
         DIABETES = case_match(DIABETES, 
                              0 ~ "No",
                              1 ~ "Yes"),
         BPMEDS = case_match(BPMEDS, 
                              0 ~ "No",
                              1 ~ "Yes"),
         PREVCHD = case_match(PREVCHD, 
                              0 ~ "No",
                              1 ~ "Yes"),
         STROKE = ifelse(((TIMESTRK < 3653) & (STROKE == 1)), "Yes", "No")
         ) %>%
  select(-TIMESTRK) %>%
  rename(Sex = SEX,
         Age = AGE,
         "Total Cholesterol" = TOTCHOL,
         "Systolic Blood Pressure" = SYSBP,
         "Diastolic Blood Pressure" = DIABP,
         "Current Smoker" = CURSMOKE,
         "Diabetes" = DIABETES,
         "Blood Pressure Medication Use" = BPMEDS,
         "Coronary Heart Disease" = PREVCHD,
         "Had a Stroke During Study" = STROKE)


dfTableOne %>% tbl_summary(by="Sex") %>%
    as_kable_extra(booktabs = TRUE, caption="Summary of patient demographics.") %>%
  kableExtra::kable_styling(font_size = 10, latex_options = "HOLD_position")
```

```{r maleDataDiagnostics, echo=FALSE, eval=FALSE}
plot(survfit(Surv(TIMESTRK, STROKE) ~  1, data = dfCardioMale))

plot(survfit(Surv(TIMESTRK, STROKE) ~  DIABETES, data = dfCardioMale))

surv_diabetes <- survfit(Surv(TIMESTRK, STROKE) ~  DIABETES, data = dfCardioMale)

ggsurvplot(surv_diabetes, conf.int=F, censor=F,
fun="cloglog")

surv_smoker <- survfit(Surv(TIMESTRK, STROKE) ~  CURSMOKE, data = dfCardioMale)

ggsurvplot(surv_smoker, conf.int=F, censor=F,
fun="cloglog")

surv_bp <- survfit(Surv(TIMESTRK, STROKE) ~  BPMEDS, data = dfCardioMale)

ggsurvplot(surv_bp, conf.int=F, censor=F,
fun="cloglog")

surv_chd <- survfit(Surv(TIMESTRK, STROKE) ~  PREVCHD, data = dfCardioMale)

ggsurvplot(surv_chd, conf.int=F, censor=F,
fun="cloglog")
```

```{r maleModel, echo=FALSE, results='hide'}
full_model_male <- coxph(Surv(TIMESTRK, STROKE) ~  AGE + DIABETES + SYSBP + PREVCHD + BPMEDS + CURSMOKE + TOTCHOL + BMI, data = dfCardioMale)

selected_model_male <- step(full_model_male, scope=list(lower=Surv(TIMESTRK, STROKE) ~ AGE + DIABETES + SYSBP, upper=full_model_male), direction="backward")

summary(selected_model_male)

summary(selected_model_male)$coef

exp(confint(selected_model_male))
```
```{r maleModelEval, echo=FALSE, eval=FALSE}
full_model_male_res <-cox.zph(full_model_male)
ggcoxzph(full_model_male_res, se=F, var="AGE")
ggcoxzph(full_model_male_res, se=F, var="DIABETES")
ggcoxzph(full_model_male_res, se=F, var="SYSBP")
# ggcoxzph(full_model_male_res, se=F, var="DIABP")
ggcoxzph(full_model_male_res, se=F, var="PREVCHD")
ggcoxzph(full_model_male_res, se=F, var="BPMEDS")
ggcoxzph(full_model_male_res, se=F, var="CURSMOKE")
ggcoxzph(full_model_male_res, se=F, var="TOTCHOL")
ggcoxzph(full_model_male_res, se=F, var="BMI")
```

```{r femaleDataDiagnostics, echo=FALSE, eval=FALSE}
plot(survfit(Surv(TIMESTRK, STROKE) ~  1, data = dfCardioFemale))

plot(survfit(Surv(TIMESTRK, STROKE) ~  DIABETES, data = dfCardioFemale))

surv_diabetes <- survfit(Surv(TIMESTRK, STROKE) ~  DIABETES, data = dfCardioFemale)

ggsurvplot(surv_diabetes, conf.int=F, censor=F,
fun="cloglog")

surv_smoker <- survfit(Surv(TIMESTRK, STROKE) ~  CURSMOKE, data = dfCardioFemale)

ggsurvplot(surv_smoker, conf.int=F, censor=F,
fun="cloglog")

surv_bp <- survfit(Surv(TIMESTRK, STROKE) ~  BPMEDS, data = dfCardioFemale)

ggsurvplot(surv_bp, conf.int=F, censor=F,
fun="cloglog")

surv_chd <- survfit(Surv(TIMESTRK, STROKE) ~  PREVCHD, data = dfCardioFemale)

ggsurvplot(surv_chd, conf.int=F, censor=F,
fun="cloglog")
```

```{r femaleModel, echo=FALSE, results='hide'}
full_model_female <- coxph(Surv(TIMESTRK, STROKE) ~  AGE + DIABETES + SYSBP + PREVCHD + BPMEDS + CURSMOKE + TOTCHOL + BMI, data = dfCardioFemale)

selected_model_female <- step(full_model_female, scope=list(lower=Surv(TIMESTRK, STROKE) ~ AGE + DIABETES + SYSBP + CURSMOKE, upper=full_model_female), direction="backward")

summary(selected_model_female)

summary(selected_model_female)$coef

exp(confint(selected_model_female))
1-exp(confint(selected_model_female))
```

```{r femaleModelEval, echo=FALSE, eval=FALSE}
full_model_female_res <-cox.zph(full_model_female)
ggcoxzph(full_model_female_res, se=F, var="AGE")
ggcoxzph(full_model_female_res, se=F, var="DIABETES")
ggcoxzph(full_model_female_res, se=F, var="SYSBP")
# ggcoxzph(full_model_female_res, se=F, var="DIABP")
ggcoxzph(full_model_female_res, se=F, var="PREVCHD")
ggcoxzph(full_model_female_res, se=F, var="BPMEDS")
ggcoxzph(full_model_female_res, se=F, var="CURSMOKE")
ggcoxzph(full_model_female_res, se=F, var="TOTCHOL")
ggcoxzph(full_model_female_res, se=F, var="BMI")
```

### Predictors of Stroke

In males, after performing variable selection, only smoking status was retained outside of the three known factors of age, diabetes, and systolic blood pressure. All three known factors were found to be significant in this study as well. In females, no variables were kept outside of those three factors, but to standardize the interpretations of the male and female models, smoking status was added to the female model. In females, only age and systolic blood were found to be significant.

In males, for each year older at baseline, the hazard of stroke increased by 6.6% (p < 0.001, 95% CI: 2.7%, 10.7%). Males with diabetes had a 380.7% higher hazard of stroke (p < 0.001, 95% CI: 109.9%, 1000.4%), and for each increase of 1 mmHg in systolic blood pressure at baseline, the hazard of stroke increases by 3.4% (p < 0.001, 95% CI: 2.2%, 4.6%). Males who smoked at baseline had a 92.4% higher hazard of stroke (p = 0.038, 95% CI: 3.6%, 257.3%).

In females, for each year older at baseline, the hazard of stroke increased by 7.8% (p < 0.001, 95% CI: 3.7%, 12.1%). Female with diabetes had a nonsignificant 81.1% higher hazard (p = 0.254, 95% CI: 34.8% lower, 403.1% higher), and for each increase of 1 mmHg in systolic blood pressure at baseline, the hazard of stroke increased by 2.6% (p < 0.001, 95% CI: 1.8%, 3.5%). Females who smoked at baseline had a nonsignificant 43.5% higher hazard of stroke (p = 0.207, 95% CI: 18.1% lower, 151.6% higher).

### Risk profiles

The probability of having a stroke within 10 years for selected risk profiles is given in table 2. The highest probability was in individuals with high blood pressure and diabetes, and the lowest probability was for average individuals, both in males and in females.

```{r maleTrajectories, echo=FALSE, eval=FALSE}
averageMale <- data.frame(AGE=c(40, 50, 60), DIABETES=0, SYSBP=mean(dfCardioMale$SYSBP),
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioMale$BMI))

highBPMale <-data.frame(AGE=c(40, 50, 60), DIABETES=0, SYSBP=160,
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioMale$BMI))

diabeticMale <-data.frame(AGE=c(40, 50, 60), DIABETES=1, SYSBP=mean(dfCardioMale$SYSBP),
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioMale$BMI))

highBPDiabeticMale <-data.frame(AGE=c(40, 50, 60), DIABETES=1, SYSBP=160,
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioMale$BMI))

smokerMale <- data.frame(AGE=c(40, 50, 60), DIABETES=0, SYSBP=mean(dfCardioMale$SYSBP),
PREVCHD=1, BPMEDS=0, CURSMOKE=1, BMI=mean(dfCardioMale$BMI))

ggsurvplot(survfit(selected_model_male, newdata = averageMale), data=dfCardioMale,censor=F,
conf.int = F,legend="right",
xlim=c(0, max(survfit(selected_model_male)$time)),
legend.labs=c("Age 40", "Age 50", "Age 60"),
title="Estimated Survival for Average Male by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5))
)

ggsurvplot(survfit(selected_model_male, newdata = highBPMale), data=dfCardioMale,censor=F,
conf.int = F,legend="right",
xlim=c(0, max(survfit(selected_model_male)$time)),
legend.labs=c("Age 40", "Age 50", "Age 60"),
title="Estimated Survival for Male with High Blood Pressure by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5))
)

ggsurvplot(survfit(selected_model_male, newdata = diabeticMale), data=dfCardioMale,censor=F,
conf.int = F,legend="right",
xlim=c(0, max(survfit(selected_model_male)$time)),
legend.labs=c("Age 40", "Age 50", "Age 60"),
title="Estimated Survival for Male with Diabetes by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5))
)

ggsurvplot(survfit(selected_model_male, newdata = highBPDiabeticMale), data=dfCardioMale,censor=F,
conf.int = F,legend="right",
xlim=c(0, max(survfit(selected_model_male)$time)),
legend.labs=c("Age 40", "Age 50", "Age 60"),
title="Estimated Survival for Male with High Blood Pressure and Diabetes by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5))
)

ggsurvplot(survfit(selected_model_male, newdata = smokerMale), data=dfCardioMale,censor=F,
conf.int = F,legend="right",
xlim=c(0, max(survfit(selected_model_male)$time)),
legend.labs=c("Age 40", "Age 50", "Age 60"),
title="Estimated Survival for Male Who Smokes by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5))
)

```

```{r femaleTrajectories, echo=FALSE, eval=FALSE}
averageFemale <-data.frame(AGE=c(40, 50, 60), DIABETES=0, SYSBP=mean(dfCardioFemale$SYSBP),
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioFemale$BMI), TOTCHOL=mean(dfCardioFemale$TOTCHOL), group="Average Female")
averageFemaleFit <- survfit(selected_model_female, newdata = averageFemale)

highBPFemale <-data.frame(AGE=c(40, 50, 60), DIABETES=0, SYSBP=160,
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioFemale$BMI), TOTCHOL=mean(dfCardioFemale$TOTCHOL), group="Female with High Blood Pressure")
highBPFemaleFit <- survfit(selected_model_female, newdata = highBPFemale)

diabeticFemale <-data.frame(AGE=c(40, 50, 60), DIABETES=1, SYSBP=mean(dfCardioFemale$SYSBP),
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioFemale$BMI), TOTCHOL=mean(dfCardioFemale$TOTCHOL), group="Female with Diabetes")
diabeticFemaleFit <- survfit(selected_model_female, newdata = diabeticFemale)

highBPDiabeticFemale <-data.frame(AGE=c(40, 50, 60), DIABETES=1, SYSBP=160,
PREVCHD=0, BPMEDS=0, CURSMOKE=0, BMI=mean(dfCardioFemale$BMI), TOTCHOL=mean(dfCardioFemale$TOTCHOL), group="Female with High Blood Pressure and Diabetes")
highBPDiabeticFemaleFit <- survfit(selected_model_female, newdata = highBPDiabeticFemale)

smokerFemale <- data.frame(AGE=c(40, 50, 60), DIABETES=0, SYSBP=mean(dfCardioFemale$SYSBP),
PREVCHD=1, BPMEDS=0, CURSMOKE=1, BMI=mean(dfCardioFemale$BMI), group="Female Who Smokes")
smokerFemaleFit <- survfit(selected_model_female, newdata = smokerFemale)

femaleProfiles <- averageFemale %>% bind_rows(highBPFemale, diabeticFemale, highBPDiabeticFemale, smokerFemale)

averageFemalePlot <- ggsurvplot(survfit(selected_model_female, newdata = averageFemale), data=dfCardioFemale,censor=F,
conf.int = F,legend="none",
xlim=c(0, max(survfit(selected_model_female)$time)),
legend.labs=NULL,
title="Estimated Survival for Average Female by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
)

highBPFemalePlot <- ggsurvplot(survfit(selected_model_female, newdata = highBPFemale), data=dfCardioFemale,censor=F,
conf.int = F,legend="none",
xlim=c(0, max(survfit(selected_model_female)$time)),
legend.labs=NULL,
title="Estimated Survival for Female with High Blood Pressure by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
)

diabeticFemalePlot <- ggsurvplot(survfit(selected_model_female, newdata = diabeticFemale), data=dfCardioFemale,censor=F,
conf.int = F,legend="right",
xlim=c(0, max(survfit(selected_model_female)$time)),
legend.labs=NULL,
title="Estimated Survival for Female with Diabetes by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
)

highBPDiabeticFemalePlot <- ggsurvplot(survfit(selected_model_female, newdata = highBPDiabeticFemale), data=dfCardioFemale,censor=F,
conf.int = F,legend="none",
xlim=c(0, max(survfit(selected_model_female)$time)),
legend.labs=NULL,
title="Estimated Survival for Female with High Blood Pressure and Diabetes by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5), legend.position = NULL)
)

smokerFemalePlot <- ggsurvplot(survfit(selected_model_female, newdata = smokerFemale), data=dfCardioFemale,censor=F,
conf.int = F,legend="none",
xlim=c(0, max(survfit(selected_model_female)$time)),
legend.labs=c("Age 40", "Age 50", "Age 60"),
title="Estimated Survival for Female Who Smokes by Age",
ggtheme=theme(plot.title = element_text(hjust = 0.5))
)
 

arrange_ggsurvplots(list(averageFemalePlot, highBPFemalePlot, diabeticFemalePlot, highBPDiabeticFemalePlot, smokerFemalePlot), nrow=3, ncol=2)
```

### Changes in risk factors

Changes in covariates are summarized in table 2. 

```{r covariates, echo=FALSE}
dfCardioWide <- dfCardio %>% filter(TIMESTRK != 0) %>%
  select(RANDID, PERIOD, AGE, DIABETES, SYSBP, DIABP, PREVCHD, BPMEDS, CURSMOKE, TOTCHOL, BMI) %>%
  pivot_wider(id_cols=RANDID, names_from = PERIOD, values_from = c(AGE, DIABETES, SYSBP, DIABP, PREVCHD, BPMEDS, CURSMOKE, TOTCHOL, BMI)) %>%
  filter(!is.na(BPMEDS_1)) %>% 
  filter(!is.na(TOTCHOL_1)) %>%
  filter(!is.na(BMI_1)) %>%
  mutate(BMIDIFF_1=abs(BMI_2-BMI_1),
         BMIDIFF_2=abs(BMI_3-BMI_2),
         SYSBPDIFF_1=abs(SYSBP_2-SYSBP_1),
         SYSBPDIFF_2=abs(SYSBP_3-SYSBP_2),
         TOTCHOLDIFF_1=abs(TOTCHOL_2-TOTCHOL_1),
         TOTCHOLDIFF_2=abs(TOTCHOL_3-TOTCHOL_2),
         DIABETESDIFF_1=ifelse(DIABETES_1==DIABETES_2, 0, 1),
         DIABETESDIFF_2=ifelse(DIABETES_3==DIABETES_2, 0, 1),
         PREVCHDDIFF_1=ifelse(PREVCHD_1==PREVCHD_2, 0, 1),
         PREVCHDDIFF_2=ifelse(PREVCHD_2==PREVCHD_3, 0, 1),
         BPMEDSDIFF_1=ifelse(BPMEDS_1==BPMEDS_2, 0, 1),
         BPMEDSDIFF_2=ifelse(BPMEDS_2==BPMEDS_3, 0, 1),
         CURSMOKEDIFF_1=ifelse(CURSMOKE_1==CURSMOKE_2, 0, 1),
         CURSMOKEDIFF_2=ifelse(CURSMOKE_2==CURSMOKE_3, 0, 1))


dfChangeTable <- dfCardioWide %>% select(BMIDIFF_1, BMIDIFF_2, SYSBPDIFF_1, SYSBPDIFF_2, TOTCHOLDIFF_1, TOTCHOLDIFF_2, DIABETESDIFF_1, DIABETESDIFF_2, PREVCHDDIFF_1, PREVCHDDIFF_2, BPMEDSDIFF_1, BPMEDSDIFF_2, CURSMOKEDIFF_1, CURSMOKEDIFF_2) %>%
  # pivot_longer(cols=c(BMIDIFF_1, BMIDIFF_2), names_to="PERIOD", names_prefix="BMIDIFF_", values_to="BMI") %>%
  pivot_longer(cols=c(BMIDIFF_1, BMIDIFF_2, SYSBPDIFF_1, SYSBPDIFF_2, TOTCHOLDIFF_1, TOTCHOLDIFF_2, DIABETESDIFF_1, DIABETESDIFF_2, PREVCHDDIFF_1, PREVCHDDIFF_2, BPMEDSDIFF_1, BPMEDSDIFF_2, CURSMOKEDIFF_1, CURSMOKEDIFF_2), names_to=c(".value", "PERIOD"), names_pattern="(.*)_(.)", names_repair="minimal", values_to=.value) %>%
  drop_na() %>%
  rename(BMI = BMIDIFF,
         "Systolic Blood Pressure" = SYSBPDIFF,
         "Total Cholesterol" = TOTCHOLDIFF,
         "Diabetes" = DIABETESDIFF,
         "Coronary Heart Disease" = PREVCHDDIFF,
         "Blood Pressure Medication Use" = BPMEDSDIFF,
         "Current Smoker" = CURSMOKEDIFF) %>%
  mutate(PERIOD = ifelse(PERIOD == 1, "Difference between\nfirst and second visit", "Difference between\nsecond and third visit"))

dfChangeTable %>% tbl_summary(by=PERIOD) %>%
  # modify_header(1 = "Difference between first and second visit", 2 = "Difference between second and third visit") %>%
  as_kable_extra(booktabs = TRUE, caption="Summary of differences in covariates between visits.") %>%
  kableExtra::kable_styling(font_size = 10, latex_options = "HOLD_position")
```

JUST NEED SYSTOLIC BP AND DIABETES ACTUALLY

## Discussion

There were a limited number of strokes present in the dataset, likely because the mean age of the participants was not that high at the start of the study and the follow up of 10 years was not enough to capture a high number of strokes.

## Reproducibility

Github Repository: https://github.com/isabellemeredith/BIOS6624

Data Directory: /Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project3/DataRaw/frmgham2.csv
