---
author: "Isabelle Meredith"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(powertools)
library(tidyr)
library(kableExtra)
```

## Project 2 Data Analysis Plan

### Data cleaning and quality control
As we are collecting longitudinal data, we expect some losses to follow up. Sample sizes are based on the number of participants who completed the study. We will not include any participants who did not have both time points in analysis. We will compare the baseline outcome values of participants who did and did not complete the study to ensure that data is missing at random, and if it is not, we will consider more advanced analysis. We will also check for outliers and data entry errors and will consider multiple imputation depending on the levels of missingness.

### Aim 1 Analytical Plan
We hypothesize that higher levels of inflammation at baseline, as indicated by cytokine and chemokine levels in circulating plasma, will predict declines in clinical memory measures and decreases in cortical thickness, another marker of memory function. Furthermore, greater increases in cytokines and chemokine levels between year 0 and year 1 will be associated with greater declines in memory measures and cortical thickness. To test for association between baseline inflammation levels and memory decline, we will use linear models between the measured cytokine and chemokines: MCP-1, Eotaxin-1, IL-6, and TNF-alpha, and the change in each of the five measures of memory between year 0 and year 1 (table 3 Cognitive Outcome Variables) as well as cortical thickness, with a separate model for each combination of predictor and outcome. We will consider baseline outcome measure, age, sex, APOE genotype, clinical dementia rating, and geriatric depression measure as precision variables, along with immunological health conditions and cardiovascular risk as potential confounders. An initial model will be fit with all variables and then backward selection will be performed to remove nonsignificant covariates. The same outcomes and covariates  will be used in another set of models with the change in each cytokine and chemokine measure between year 0 and 1 as the predictor instead to test hypothesis 1b that increases in cytokine and chemokine levels result in declines in memory function. We will use an empirical correction of $\alpha = 0.005$ for multiple comparisons, as we expect that our predictors and outcomes will be correlated, and therefore we would expect most or all relationships to be significant if a non-spurious relationship exists. This allows us to assess the relationships without needing a strict correction. 

We performed power calculations using a one sample correlation test as a simplification of the linear models we will perform for each outcome and predictor. Calculations were made with the PowerTools package in R. We estimated power based on the range of correlations in preliminary data for 30 subjects for the IL6 and MCP1 cytokines and the CVLT and cortical thickness measures. Due to the correlation among predictors and outcomes, we expect to see a similar range of correlations for all combinations, so the power calculations are applicable to all models. As we have both precision and confounding variables, we will leave the power estimates unadjusted rather than making conservative or unconservative estimates. Estimations use a sample size of 175 based on reasonable recruitment levels at participating clinics when accounting for potential 10% attrition. We achieve 80% power at a correlation of 0.27, which was at the low end of the preliminary correlations, which were between 0.26 and 0.69. Furthermore, we maintain at least 70% power for the range of correlations found in the preliminary data (table 1). 

```{r, echo=FALSE, results='hide'}
dfPrelim <- read.csv("/Users/isabelle/Documents/School/5 Masters Year 2/Advanced Data Analysis/BIOS6624/Project2/PrelimData.csv")

cor(dfPrelim$CVLT_CNG3, dfPrelim$IL_6)
cor(dfPrelim$CVLT_CNG3, dfPrelim$MCP_1)
cor(dfPrelim$CORT_CNG3, dfPrelim$IL_6)
cor(dfPrelim$CORT_CNG3, dfPrelim$MCP_1)

# Aim 1
corr.1samp(N=175, alpha=0.005, rho0=0, sides=2, v=FALSE, power=NULL, rhoA = 0.27)

correlations <- seq(0.25, 0.70, by=0.05)
powers <- lapply(correlations, corr.1samp, N=175, alpha=0.005, rho0=0, sides=2, v=FALSE, power=NULL)
dfPowerAim1 <- data.frame(correlation = correlations)
dfPowerAim1$powers <- as.numeric(powers)
```

```{r, echo=FALSE}
dfPowersWide <- data.frame(t(dfPowerAim1[,"powers"]))
colnames(dfPowersWide) <- dfPowerAim1$correlation
rownames(dfPowersWide) <- c("Power")
dfPowersWide <- dfPowersWide

dfPowersWide %>% kable(digits=3, col.names = c("Correlation", dfPowerAim1$correlation), caption="Estimated power to detect the given correlation for N=175.")
```

### Aim 1 Alternative Approach
Although study drop out is expected to be missing at random, if we find that there are trends in missingness, we will consider an alternative approach that also models the missingness process. 

### Aim 2 Analytical Plan
We hypothesize that there will be an interaction between inflammation and amyloid deposition such that declines in memory function and cortical thickness will decline faster with higher levels of both inflammation and amyloid deposition compared to the decline as either increases alone. To test for this association, we will use general linear models with the change between year 0 and year 1 for each of our outcomes listed in aim 1. Change in cytokine and chemokine levels, baseline amyloid deposition, and the interaction between the two will be the predictors, with a separate model for each predictor. We will adjust for the potential covariates listed in aim 1 with backward selection and we will use the same empirical multiple comparisons correction of $\alpha = 0.005$  

We performed power calculations using a two sample correlation test as a simplification of the interaction when considering amyloid deposition as a binary variable. The correlations were considered to be the correlation for the inflammation with and without significant amyloid deposition. Calculations were made with the PowerTools package in R. As in aim 1, we use the preliminary correlations and we do not use conservative or unconservative estimates. We also retain the sample size of 175. Powers were calculated across varying potential percentages of subjects with significant amyloid deposition: 50% (table 2), 33% (table 3), 25% (table 4). When the initial correlation is at the low end of our preliminary correlations, we need a large difference in correlation of 0.45 with the amyloid deposition to achieve 80% power, but at higher initial correlations, a difference in correlation of 0.2-0.3 is detectable with 80% power, though power decreases with a lower portion of subjects with amyloid deposition. 

### Aim 2 Alternative Approach
As in aim 1, if missingness is not at random, we will consider an alternative approach that also models the missingness process. Additionally, although the recruitment methodology is likely to capture many people with amyloid deposition, we are underpowered if the percentage of subjects is below 25%. If we find that amyloid deposition rates are below 25%, we will not use the multiple testing comparison in order to ensure adequate power (table 5) and will instead rely on internal consistency across the correlated outcomes to check the validity of our findings with more cautious interpretation.

```{r, echo=FALSE}
correlation_calc <- function(y, n1, n.ratio, alpha=0.005) {if (y['Var1']+y['Var2'] < 1) corr.2samp(y['Var1'],min(y['Var1']+y['Var2'], 0.99), alpha=alpha, sides=2, v=FALSE, power=NULL, n1=n1, n.ratio=n.ratio) else NA}

options(knitr.kable.NA = 'N/A')

# Aim 2
correlation1 <- seq(0.2, 0.7, by=0.1)
correlation2 <- seq(0.2, 0.45, by=0.05)
nCol <- length(correlation2)
correlationGrid <- expand.grid(correlation1, correlation2)
powers <- matrix(apply(correlationGrid, MARGIN=1, correlation_calc, n1=87.5, n.ratio=1), nrow=6, ncol=nCol)
dfPowerAim2Prop50 <- as.data.frame(powers)
colnames(dfPowerAim2Prop50) <- correlation2
rownames(dfPowerAim2Prop50) <- correlation1
dfPowerAim2Prop50 %>% kable(col.names = c("Initial Correlation", correlation2), digits=3, caption="Estimated power to detect a difference in correlation for N=175 with significant amyloid deposition for fifty percent of subjects using alpha of 0.005.") %>% 
  add_header_above(header = c(" " = 1, "Difference in Correlation" = nCol))

correlationGrid <- expand.grid(correlation1, correlation2)
powers <- matrix(apply(correlationGrid, MARGIN=1, correlation_calc, n1=2*175/3, n.ratio=1/2), nrow=6, ncol=nCol)
dfPowerAim2Prop33 <- as.data.frame(powers)
colnames(dfPowerAim2Prop33) <- correlation2
rownames(dfPowerAim2Prop33) <- correlation1
dfPowerAim2Prop33 %>% kable(col.names = c("Initial Correlation", correlation2), digits=3, caption="Estimated power to detect a difference in correlation for N=175 with significant amyloid deposition for one third of the subjects using alpha of 0.005.") %>% 
  add_header_above(header = c(" " = 1, "Difference in Correlation" = nCol))

correlationGrid <- expand.grid(correlation1, correlation2)
powers <- matrix(apply(correlationGrid, MARGIN=1, correlation_calc, n1=3*175/4, n.ratio=1/3), nrow=6, ncol=nCol)
dfPowerAim2Prop25 <- as.data.frame(powers)
colnames(dfPowerAim2Prop25) <- correlation2
rownames(dfPowerAim2Prop25) <- correlation1
dfPowerAim2Prop25 %>% kable(col.names = c("Initial Correlation", correlation2), digits=3, caption="Estimated power to detect a difference in correlation for N=175 with significant amyloid deposition for twenty five percent of subjects using alpha of 0.005.") %>% 
  add_header_above(header = c(" " = 1, "Difference in Correlation" = nCol))

correlationGrid <- expand.grid(correlation1, correlation2)
powers <- matrix(apply(correlationGrid, MARGIN=1, correlation_calc, n1=4*175/5, n.ratio=1/4, alpha=0.05), nrow=6, ncol=nCol)
dfPowerAim2Prop20 <- as.data.frame(powers)
colnames(dfPowerAim2Prop20) <- correlation2
rownames(dfPowerAim2Prop20) <- correlation1
dfPowerAim2Prop20 %>% kable(col.names = c("Initial Correlation", correlation2), digits=3, caption="Estimated power to detect a difference in correlation for N=175 with significant amyloid deposition for twenty percent of subjects using alpha of 0.05.") %>% 
  add_header_above(header = c(" " = 1, "Difference in Correlation" = nCol))
```