---
title: "Analysis"
author: "Isabelle Meredith"
date: "2026-01-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(lme4)
library(lmerTest)
library(ggplot2)
library(tidyr)
```

## R Markdown

use nmol/L for cortisol levels

within 7.5 minutes is good
within 15 minutes is adequate

Time trend spikes 30 minutes after waking and then declines

Analysis plan

what is the agreement between the subject’s recordings of sampling times compared to the times recorded by an electronic monitoring cap?
Scatterplot of recorded time vs electronic time
Analyze correlation?
Test if mean difference is significantly different from 0?
Mixed effects model with subject id?

are subjects adhering accurately to the +30min and +10 hour sampling times required by a study protocol?
proportion within the time?

what are the changes of cortisol and DHEA over time?
nonlinear or piecewise linear trend


Question 1)
linear mixed effects model between the two measures with random intercept for subject. if intercept is 0, no bias. if slope is 1, correlation is perfect

book ~ cap + (1 | subj)
use minutes from waking, as calculated by us (not the preexisting variable)

Question 2)
+= 7.5 minutes
+= 15 minutes

Can use descriptive statistics and summarize for this. what is the mean and sd of difference. what proportion is within each of these intervals

could use confidence interval for proportion if we want

Question 3)
Piecewise mixed effects model?
Break at 30 minutes after waking

table 1
book time - cap time
dhea and cortisol overall and by time point

```{r}
dfSleep <- read.csv("./DataRaw/Project0_Clean_v2.csv")

dfSleep <- dfSleep %>% rename(BookletMinutes = Booklet..Sample.interval.Decimal.Time..mins.,
                   MEMsMinutes = MEMs..Sample.interval.Decimal.Time..mins.)
```

Question 1
```{r}
dfSleep <- dfSleep %>% separate_wider_delim(Booket..Clock.Time, delim=":", names=c("HoursBooklet", "MinutesBooklet"), too_few="align_start", cols_remove = FALSE) %>% 
  mutate(
    HoursBooklet = as.numeric(HoursBooklet),
    MinutesBooklet = as.numeric(MinutesBooklet),
    TimeSinceMidnightBooklet = ifelse(is.na(HoursBooklet) | is.na(MinutesBooklet), NA, HoursBooklet*60 + MinutesBooklet)) %>% separate_wider_delim(MEMs..Clock.Time, delim=":", names=c("HoursMEMs", "MinutesMEMs"), too_few="align_start", cols_remove = FALSE) %>% 
  mutate(
    HoursMEMs = as.numeric(HoursMEMs),
    MinutesMEMs = as.numeric(MinutesMEMs),
    TimeSinceMidnightMEMs = ifelse(is.na(HoursMEMs) | is.na(MinutesMEMs), NA, HoursMEMs*60 + MinutesMEMs)) 

dfSleep %>% ggplot(aes(x=TimeSinceMidnightMEMs, y=TimeSinceMidnightBooklet)) +
  geom_point() +
  geom_abline(intercept=0, slope=1)

sum(is.na(dfSleep$TimeSinceMidnightBooklet) | is.na(dfSleep$TimeSinceMidnightMEMs))
sum(is.na(dfSleep$Cortisol..nmol.L.) & (is.na(dfSleep$TimeSinceMidnightBooklet) | is.na(dfSleep$TimeSinceMidnightMEMs)))
sum(is.na(dfSleep$DHEA..nmol.L.) & (is.na(dfSleep$TimeSinceMidnightBooklet) | is.na(dfSleep$TimeSinceMidnightMEMs)))
```

```{r}
associationModel <- lmer(TimeSinceMidnightBooklet ~ TimeSinceMidnightMEMs + (1 | SubjectID), data=dfSleep)
summary(associationModel)
```
Question 2
literally just get the proportions that are within 7.5 and 15 for the booklet and cap. do a confidence interval if I'm feeling bold
```{r}
dfSleep <- dfSleep %>% mutate(
  BookletWithin7.5 = case_when(
    Collection.Sample == 2 ~ abs(BookletMinutes - 30) <= 7.5,
    Collection.Sample == 4 ~ abs(BookletMinutes - 600) <= 7.5,
  ),
  BookletWithin15 = case_when(
    Collection.Sample == 2 ~ abs(BookletMinutes - 30) <= 15,
    Collection.Sample == 4 ~ abs(BookletMinutes - 600) <= 15,
  ),
  MEMsWithin7.5 = case_when(
    Collection.Sample == 2 ~ abs(MEMsMinutes - 30) <= 7.5,
    Collection.Sample == 4 ~ abs(MEMsMinutes - 600) <= 7.5,
  ),
  MEMsWithin15 = case_when(
    Collection.Sample == 2 ~ abs(MEMsMinutes - 30) <= 15,
    Collection.Sample == 4 ~ abs(MEMsMinutes - 600) <= 15,
  ))






tab <- matrix(c(sum(dfSleep$BookletWithin15, na.rm=TRUE) / sum(!is.na(dfSleep$BookletWithin15)), sum(dfSleep$BookletWithin7.5, na.rm=TRUE) / sum(!is.na(dfSleep$BookletWithin7.5)), sum(dfSleep$MEMsWithin15, na.rm=TRUE) / sum(!is.na(dfSleep$MEMsWithin15)), sum(dfSleep$MEMsWithin7.5, na.rm=TRUE) / sum(!is.na(dfSleep$MEMsWithin7.5))), ncol=2, byrow=TRUE)
colnames(tab) <- c('Within 7.5 Minutes','Within 15 Minutes')
rownames(tab) <- c('Booklet','Cap')
tab <- as.table(tab)
print(tab)
```

Question 3
Use booklet time
break point at 30 minutes

```{r}
hist(dfSleep$DHEA..nmol.L.)
hist(log(dfSleep$DHEA..nmol.L.))


```